---
title: "hawk2"
author: "Edgardo L. Rosado Ramos"
date: "2025-09-16"
output: html_document
---

```{r phyloseq-setup, message=FALSE, warning=FALSE}


library(phyloseq)
library(dplyr)
library(tidyr)
library(tibble)


otu_file <- "final.opti_mcc.0.03.subsample.shared"
tax_file <- "final.opti_mcc.0.03.cons.taxonomy"
metadata_file <- "hawkm.design.txt"


otu_df <- read.table(otu_file, header = TRUE, sep = "\t")
rownames(otu_df) <- otu_df$Group
otu_numeric_df <- otu_df[, -(1:3)]
otu_matrix <- t(otu_numeric_df)



tax_df_raw <- read.table(tax_file, header = TRUE, sep = "\t")
tax_df_separated <- tax_df_raw %>%
  separate(
    col = Taxonomy, 
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus"), 
    sep = ";",
    extra = "warn",
    fill = "right"
  )
tax_df_almost_clean <- select(tax_df_separated, -7)
tax_df_clean <- tax_df_almost_clean %>%
   mutate(across(where(is.character), ~gsub("\\(\\d+\\)|k__|p__|c__|o__|f__|g__", "", .)))
rownames(tax_df_clean) <- tax_df_clean$OTU
tax_matrix <- tax_df_clean %>%
  select(any_of(c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus"))) %>%
  as.matrix()


metadata_df <- read.table(metadata_file, header = TRUE, sep = "\t")
rownames(metadata_df) <- metadata_df$SampleID



OTU = otu_table(otu_matrix, taxa_are_rows = TRUE)
TAX = tax_table(tax_matrix)
META = sample_data(metadata_df)


valid_sample_names <- sample_names(OTU)
META_synced <- sample_data(META)[valid_sample_names, ]

otus_to_keep <- taxa_names(OTU)
TAX_synced <- tax_table(TAX)[otus_to_keep, ]

physeq_final <- phyloseq(OTU, TAX_synced, META_synced)


# --- Part 6: Verify Your Final Object ---
print("--- Final Phyloseq Object Created ---")
print(physeq_final)
# ===================================================================
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)
library(ComplexHeatmap)
library(DESeq2)
library(rstatix) 


sample_type_cols <- c("Buccal"="#1f78b4", "Cloaca"="#33a02c", "Control"="#e31a1c")
sample_type_shapes <- c("Buccal"=16, "Cloaca"=17, "Control"=15)

theme_set(theme_bw())
```

## Introduction

This report details the analysis of 16S rRNA gene sequencing data from Buccal, Cloaca, and Control samples collected from the Hawksbill Sea Turtle (*Eretmochelys imbricata*).

---

## 1. Alpha Diversity Analysis

Alpha diversity measures the species richness and evenness within a single sample. Here we compare the Shannon diversity index across our three sample types.

```{r alpha-diversity-plot}

alpha_plot <- plot_richness(physeq_final, x = "SampleType", measures = c("Shannon")) +
  geom_boxplot(aes(fill = SampleType), alpha = 0.7) +
  scale_fill_manual(values = sample_type_cols, name = "Sample Type") +
  labs(title = "Alpha Diversity (Shannon Index)", x = "Sample Type", y = "Shannon Diversity") +
  theme(legend.position = "none")

print(alpha_plot)
```

**Statistical Test:** We use the Kruskal-Wallis test to determine if the observed differences in Shannon diversity are statistically significant.

```{r alpha-diversity-stats}
alpha_data <- estimate_richness(physeq_final, measures = "Shannon")
alpha_data$SampleType <- sample_data(physeq_final)$SampleType

kruskal_test_result <- kruskal_test(data = alpha_data, Shannon ~ SampleType)

knitr::kable(kruskal_test_result, caption = "Kruskal-Wallis test for Shannon Diversity")
print(kruskal_test_result)
```

---

## 2. Beta Diversity Analysis

Beta diversity measures the difference in community composition between samples. We use Principal Coordinates Analysis (PCoA) on a Bray-Curtis dissimilarity matrix to visualize these differences.

```{r pcoa-plot}
pcoa_ord <- ordinate(physeq_final, method = "PCoA", distance = "bray")

pcoa_plot <- plot_ordination(physeq_final, pcoa_ord, type = "samples", color = "SampleType", shape = "SampleType") +
  geom_point(size = 4) +
  stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(name = "Sample Type", values = sample_type_cols) +
  scale_shape_manual(name = "Sample Type", values = sample_type_shapes) +
  labs(title = "PCoA of E. imbricata Microbial Communities",
       subtitle = "Bray-Curtis Dissimilarity")

print(pcoa_plot)
```

**Statistical Test:** We use a PERMANOVA test (`adonis2`) to determine if the clustering of sample groups is statistically significant.

```{r beta-diversity-stats, results='asis'}
bray_dist <- phyloseq::distance(physeq_final, method = "bray")
sample_df <- data.frame(sample_data(physeq_final))

permanova_result <- adonis2(bray_dist ~ SampleType, data = sample_df, permutations = 999)

knitr::kable(permanova_result, caption = "PERMANOVA results for Beta Diversity")
print(permanova_result)
```

---

## 3. Taxonomic Composition

### Relative Abundance Bar Plot
This plot shows the relative abundance of the major bacterial Phyla across each sample.

```{r bar-plot}
physeq_phylum <- tax_glom(physeq_final, taxrank = "Phylum")

bar_plot <- plot_bar(physeq_phylum, x = "SampleID", fill = "Phylum") +
  facet_wrap(~SampleType, scales = "free_x", nrow = 1) +
  labs(title = "Relative Abundance by Phylum", x = "Sample", y = "Relative Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8))

print(bar_plot)
```

### Heatmap of Abundant Genera
This heatmap shows the relative abundance of the top 30 most abundant genera.

```{r heatmap}
physeq_genus <- tax_glom(physeq_final, taxrank = "Genus", NArm = FALSE)

top30_taxa <- names(sort(taxa_sums(physeq_genus), decreasing = TRUE)[1:30])
physeq_top30 <- prune_taxa(top30_taxa, physeq_genus)

plot_heatmap(physeq_top30, 
             sample.label = "SampleType", 
             taxa.label = "Genus", 
             sample.order = "SampleType",
             low = "#FFFFFF", high = "#0000FF", na.value = "white",
             title = "Heatmap of Top 30 Most Abundant Genera")
```

---

## 4. Differential Abundance Analysis (DESeq2)
To identify which specific taxa are significantly different between groups, we use DESeq2.
**Note:** DESeq2 should ideally be run on non-rarefied data. This example uses your rarefied data for demonstration.

```{r deseq2-analysis}
# Add 1 to all counts to handle zeros, a requirement for DESeq2
physeq_deseq <- physeq_final
otu_table(physeq_deseq) <- otu_table(physeq_deseq) + 1

# Convert phyloseq object to DESeq2 object
deseq_obj <- phyloseq_to_deseq2(physeq_deseq, design = ~ SampleType)

# Run DESeq2
deseq_results <- DESeq(deseq_obj, test = "Wald", fitType = "parametric")

# --- Compare Buccal vs. Control ---
res_buccal_vs_control <- results(deseq_results, contrast = c("SampleType", "Buccal", "Control"))
res_buccal_vs_control_sig <- subset(res_buccal_vs_control, padj < 0.05)
res_buccal_vs_control_sig <- cbind(as.data.frame(res_buccal_vs_control_sig), as.data.frame(tax_table(physeq_final)[rownames(res_buccal_vs_control_sig), ]))

# --- Compare Cloaca vs. Control ---
res_cloaca_vs_control <- results(deseq_results, contrast = c("SampleType", "Cloaca", "Control"))
res_cloaca_vs_control_sig <- subset(res_cloaca_vs_control, padj < 0.05)
res_cloaca_vs_control_sig <- cbind(as.data.frame(res_cloaca_vs_control_sig), as.data.frame(tax_table(physeq_final)[rownames(res_cloaca_vs_control_sig), ]))

print("--- DESeq2 Analysis Completed ---")
```

### Table: Taxa Significantly Different Between Buccal and Control
```{r deseq2-table-1, results='asis'}
knitr::kable(res_buccal_vs_control_sig, caption = "Differentially abundant taxa (p-adj < 0.05) between Buccal and Control samples.")

print(res_buccal_vs_control_sig)

```

### Table: Taxa Significantly Different Between Cloaca and Control
```{r deseq2-table-2, results='asis'}
knitr::kable(res_cloaca_vs_control_sig, caption = "Differentially abundant taxa (p-adj < 0.05) between Cloaca and Control samples.")
print(res_cloaca_vs_control_sig)

```