---
title: "Feed Status Effect on Microbiome at 15 Days"
author: "Edgardo L. Rosado Ramos"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8)
```

# Analysis Overview

**Research Question:** Does feeding status (Fed vs. Unfed) affect the gut microbiome composition in 15-day-old snakes?

**Hypothesis:** Fed snakes will have different microbiome composition due to nutrient influx and altered gut environment.

# Setup

```{r libraries}
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(kableExtra)
library(DESeq2)

theme_set(theme_bw(base_size = 14))
```

# Data Loading

```{r load-data}
# Load OTU Table
otu_raw <- read.table("kohler.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group 
otu_mat <- t(otu_mat)
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)

# Load Taxonomy
tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU
taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
  x <- x[nchar(x) > 0]
  x <- gsub("\\(.*?\\)|\\\"", "", x)
  trimws(x)
})
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)
tax_tab <- tax_table(as.matrix(taxonomy_mat))

# Load Metadata
Kohler.design <- read.table("Kohlerdesign.txt",
                            header = TRUE, sep = "\t", row.names = 1,
                            stringsAsFactors = TRUE, check.names = FALSE)

# Create Phyloseq
common_samples <- intersect(sample_names(otu_tab), rownames(Kohler.design))
otu_tab_pruned <- prune_samples(common_samples, otu_tab)
sample_data_obj <- sample_data(Kohler.design[common_samples, ])
physeq <- phyloseq(otu_tab_pruned, tax_tab, sample_data_obj)

# Rarefy
physeq_filt <- filter_taxa(physeq, function(x) sum(x > 3) > (0.2 * length(x)), TRUE)
min_depth <- min(sample_sums(physeq_filt))
set.seed(42)
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)

# Alpha Diversity
alpha_diversity_rarefied <- estimate_richness(physeq_rarefied, measures = c("Observed", "Shannon", "Chao1"))
sample_metadata_rarefied <- as.data.frame(sample_data(physeq_rarefied))
alpha_data_rarefied <- cbind(alpha_diversity_rarefied, sample_metadata_rarefied)
```

# Subset to 15-Day Samples

```{r subset-15days}
physeq_15days <- subset_samples(physeq_rarefied, Age == "15+days")
alpha_15days <- alpha_data_rarefied %>% filter(Age == "15+days")

cat("Total 15-day samples:", nsamples(physeq_15days), "\n")
cat("Sample breakdown:\n")
table(sample_data(physeq_15days)$Feed, sample_data(physeq_15days)$SampleType)
```

# Sample Distribution

```{r sample-table}
sample_summary <- alpha_15days %>%
  group_by(SampleType, Feed) %>%
  summarise(n = n(), .groups = "drop")

knitr::kable(sample_summary, 
             caption = "Sample Distribution at 15 Days",
             col.names = c("Treatment", "Feed Status", "Sample Size")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

# Alpha Diversity Analysis

## Shannon Diversity

```{r alpha-shannon, fig.height=6, fig.width=10}
p_shannon <- ggplot(alpha_15days, aes(x = Feed, y = Shannon, fill = Feed)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  facet_wrap(~ SampleType) +
  scale_fill_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 5) +
  labs(title = "Shannon Diversity: Fed vs. Unfed Snakes at 15 Days",
       y = "Shannon Diversity Index",
       x = "Feeding Status") +
  theme(legend.position = "top", text = element_text(size = 14))

print(p_shannon)
ggsave("FeedStatus_Shannon.pdf", p_shannon, width = 10, height = 6)
```

## Observed OTUs

```{r alpha-observed, fig.height=6, fig.width=10}
p_observed <- ggplot(alpha_15days, aes(x = Feed, y = Observed, fill = Feed)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  facet_wrap(~ SampleType) +
  scale_fill_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 5) +
  labs(title = "Species Richness: Fed vs. Unfed Snakes at 15 Days",
       y = "Observed OTUs (Species Richness)",
       x = "Feeding Status") +
  theme(legend.position = "top", text = element_text(size = 14))

print(p_observed)
ggsave("FeedStatus_Richness.pdf", p_observed, width = 10, height = 6)
```

## Summary Statistics

```{r alpha-summary}
alpha_stats <- alpha_15days %>%
  group_by(SampleType, Feed) %>%
  summarise(
    n = n(),
    Shannon_mean = mean(Shannon, na.rm = TRUE),
    Shannon_sd = sd(Shannon, na.rm = TRUE),
    Observed_mean = mean(Observed, na.rm = TRUE),
    Observed_sd = sd(Observed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Shannon = paste0(round(Shannon_mean, 2), " ± ", round(Shannon_sd, 2)),
    Richness = paste0(round(Observed_mean, 1), " ± ", round(Observed_sd, 1))
  ) %>%
  dplyr::select(SampleType, Feed, n, Shannon, Richness)

knitr::kable(alpha_stats,
             col.names = c("Treatment", "Fed?", "n", "Shannon (Mean ± SD)", "Richness (Mean ± SD)"),
             caption = "Alpha Diversity Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

write.csv(alpha_stats, "FeedStatus_AlphaStats.csv", row.names = FALSE)
```

# Beta Diversity Analysis

## PERMANOVA: Feed Status Effect

```{r permanova-feed}
# Calculate Bray-Curtis distance
bray_15days <- phyloseq::distance(physeq_15days, method = "bray")
metadata_15days <- data.frame(sample_data(physeq_15days))

# PERMANOVA testing Feed + SampleType + Mother
set.seed(42)
permanova_feed <- adonis2(
  bray_15days ~ Feed + SampleType + Mother,
  data = metadata_15days,
  permutations = 9999,
  by = "margin"
)

print(permanova_feed)

# Create effect size table
effect_table_feed <- data.frame(
  Factor = c("Feed Status", "Treatment (Muck/Sterile)", "Maternal ID"),
  R2 = c(
    permanova_feed$R2[which(rownames(permanova_feed) == "Feed")],
    permanova_feed$R2[which(rownames(permanova_feed) == "SampleType")],
    permanova_feed$R2[which(rownames(permanova_feed) == "Mother")]
  ),
  F_stat = c(
    permanova_feed$F[which(rownames(permanova_feed) == "Feed")],
    permanova_feed$F[which(rownames(permanova_feed) == "SampleType")],
    permanova_feed$F[which(rownames(permanova_feed) == "Mother")]
  ),
  p_value = c(
    permanova_feed$`Pr(>F)`[which(rownames(permanova_feed) == "Feed")],
    permanova_feed$`Pr(>F)`[which(rownames(permanova_feed) == "SampleType")],
    permanova_feed$`Pr(>F)`[which(rownames(permanova_feed) == "Mother")]
  )
) %>%
  mutate(
    `R² (%)` = round(R2 * 100, 1),
    `F-statistic` = round(F_stat, 2),
    `p-value` = format.pval(p_value, digits = 3),
    Significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  dplyr::select(Factor, `R² (%)`, `F-statistic`, `p-value`, Significance)

knitr::kable(effect_table_feed,
             caption = "PERMANOVA: Drivers of Microbiome Variation at 15 Days",
             align = "lcccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#fff3cd")  # Highlight Feed Status

write.csv(effect_table_feed, "FeedStatus_PERMANOVA.csv", row.names = FALSE)
```

## PCoA Ordination

```{r pcoa-feed, fig.height=7, fig.width=12}
pcoa_15days <- ordinate(physeq_15days, method = "PCoA", distance = bray_15days)

p_pcoa <- plot_ordination(physeq_15days, pcoa_15days, 
                          color = "Feed", shape = "SampleType") +
  geom_point(size = 6, alpha = 0.8) +
  scale_color_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60"),
                     name = "Fed?") +
  scale_shape_manual(values = c(16, 17), name = "Treatment") +
  labs(
    title = "PCoA: Microbiome Composition by Feed Status (15 Days)",
    subtitle = paste0("PERMANOVA - Feed: R² = ", 
                     round(effect_table_feed$`R² (%)`, 1)[1], 
                     "%, p = ", effect_table_feed$`p-value`[1]),
    x = paste0("PCoA1 (", round(pcoa_15days$values$Relative_eig[1] * 100, 1), "%)"),
    y = paste0("PCoA2 (", round(pcoa_15days$values$Relative_eig[2] * 100, 1), "%)")
  ) +
  theme(legend.position = "right", text = element_text(size = 14))

print(p_pcoa)
ggsave("FeedStatus_PCoA.pdf", p_pcoa, width = 12, height = 7)
```

# Differential Abundance Analysis

## DESeq2: Taxa Responsive to Feeding

```{r deseq2-feed}
# Subset original (non-rarefied) data for DESeq2
physeq_15days_raw <- subset_samples(physeq, Age == "15+days")
physeq_15days_raw <- prune_taxa(taxa_sums(physeq_15days_raw) > 0, physeq_15days_raw)

# Convert to DESeq2
ds_obj_feed <- phyloseq_to_deseq2(physeq_15days_raw, ~ SampleType + Feed)
ds_res_feed <- DESeq(ds_obj_feed, test = "Wald", fitType = "parametric")

# Get results for Fed vs. Unfed
res_feed <- results(ds_res_feed, contrast = c("Feed", "Yes", "No"))

# Filter significant taxa
sig_taxa_feed <- as.data.frame(res_feed) %>%
  filter(padj < 0.1) %>%  # Use relaxed threshold for pilot study
  arrange(desc(abs(log2FoldChange)))

if (nrow(sig_taxa_feed) > 0) {
  tax_info_feed <- as.data.frame(tax_table(physeq)[rownames(sig_taxa_feed), ])
  sig_taxa_feed_annotated <- cbind(sig_taxa_feed, tax_info_feed)
  sig_taxa_feed_annotated$Taxon <- paste(sig_taxa_feed_annotated$Family, 
                                         sig_taxa_feed_annotated$Genus, sep = " : ")
  
  cat("\n## Taxa Responsive to Feeding (FDR < 0.1):\n\n")
  knitr::kable(sig_taxa_feed_annotated[, c("log2FoldChange", "padj", "Taxon")],
               caption = "Differentially Abundant Taxa: Fed vs. Unfed",
               col.names = c("log2 FC (Fed/Unfed)", "Adjusted p-value", "Taxon"),
               digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
  
  write.csv(sig_taxa_feed_annotated, "FeedStatus_DifferentialAbundance.csv")
} else {
  cat("\nNo taxa significantly different at FDR < 0.1\n")
  cat("This may be due to small sample size. Consider relaxing threshold or qualitative interpretation.\n")
}
```

# Taxonomic Composition

## Family-Level Composition

```{r taxonomy-family, fig.height=7, fig.width=12}
physeq_family_15 <- tax_glom(physeq_15days, taxrank = "Family")
physeq_rel_f_15 <- transform_sample_counts(physeq_family_15, function(x) x / sum(x))
melted_phy_f_15 <- psmelt(physeq_rel_f_15)

top_families_15 <- names(sort(tapply(melted_phy_f_15$Abundance, melted_phy_f_15$Family, sum), 
                              decreasing = TRUE))[1:12]
melted_phy_f_15$Family_Collapsed <- ifelse(melted_phy_f_15$Family %in% top_families_15, 
                                            as.character(melted_phy_f_15$Family), "Other")

p_taxonomy <- ggplot(melted_phy_f_15, aes(x = Sample, y = Abundance, fill = Family_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(Feed ~ SampleType, scales = "free_x", space = "free") +
  scale_fill_viridis_d(name = "Family", option = "turbo") +
  labs(title = "Family-Level Composition: Feed Status Comparison (15 Days)",
       y = "Relative Abundance") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))

print(p_taxonomy)
ggsave("FeedStatus_Taxonomy.pdf", p_taxonomy, width = 12, height = 7)
```

# Summary

## Key Findings

Based on the analysis of 15-day-old snakes:

1. **Alpha Diversity:** 
   - Shannon diversity: [See plot above for statistical significance]
   - Species richness: [See plot above for statistical significance]

2. **Beta Diversity (PERMANOVA):**
   - Feed Status R²: **`r round(effect_table_feed$'R² (%)'[1], 1)`%** (p = `r effect_table_feed$'p-value'[1]`)
   - Feed effect relative to Treatment and Maternal ID shown in table above

3. **Differential Abundance:**
   - Number of responsive taxa (FDR < 0.1): **`r if(exists('sig_taxa_feed')) nrow(sig_taxa_feed) else 0`**

4. **Interpretation:**
   - Feeding status [does/does not] significantly alter microbiome composition at 15 days
   - Effect size is [small/medium/large] compared to other factors
   - Consider in context of sample size limitations

## Session Info

```{r session-info}
sessionInfo()
```
