---
title: "Feed Status Effect on Microbiome"
author: "Edgardo L. Rosado Ramos"
date: "2025-11-21"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8)
```

# Analysis Overview

**Research Question:** Does feeding status (Fed vs. Unfed) affect the gut microbiome composition?

**Statistical Issue Identified:** At 15 days, Feed status is confounded with Treatment (only 1 Muck-Fed sample).

**Revised Strategy:**
1. **Primary Analysis:** Focus on Sterile group at 15 days (balanced: 4 Fed vs. 4 Unfed)
2. **Exploratory:** Check Feed distribution across all ages to identify where Muck can be integrated

---

# Setup

```{r libraries}
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(kableExtra)
library(DESeq2)

theme_set(theme_bw(base_size = 14))
```

# Data Loading

```{r load-data}
# Load OTU Table
otu_raw <- read.table("kohler.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group 
otu_mat <- t(otu_mat)
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)

# Load Taxonomy
tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU
taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
  x <- x[nchar(x) > 0]
  x <- gsub("\\(.*?\\)|\\\"", "", x)
  trimws(x)
})
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)
tax_tab <- tax_table(as.matrix(taxonomy_mat))

# Load Metadata
Kohler.design <- read.table("Kohlerdesign.txt",
                            header = TRUE, sep = "\t", row.names = 1,
                            stringsAsFactors = TRUE, check.names = FALSE)

# Create Phyloseq
common_samples <- intersect(sample_names(otu_tab), rownames(Kohler.design))
otu_tab_pruned <- prune_samples(common_samples, otu_tab)
sample_data_obj <- sample_data(Kohler.design[common_samples, ])
physeq <- phyloseq(otu_tab_pruned, tax_tab, sample_data_obj)

# Rarefy
physeq_filt <- filter_taxa(physeq, function(x) sum(x > 3) > (0.2 * length(x)), TRUE)
min_depth <- min(sample_sums(physeq_filt))
set.seed(42)
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)

# Alpha Diversity
alpha_diversity_rarefied <- estimate_richness(physeq_rarefied, measures = c("Observed", "Shannon", "Chao1"))
sample_metadata_rarefied <- as.data.frame(sample_data(physeq_rarefied))
alpha_data_rarefied <- cbind(alpha_diversity_rarefied, sample_metadata_rarefied)
```

---

# Part 1: Feed Distribution Survey

**Goal:** Identify ages where Feed status is balanced across both Muck and Sterile treatments.

```{r feed-distribution}
# Create comprehensive table
feed_dist <- alpha_data_rarefied %>%
  group_by(Age, SampleType, Feed) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = Feed, values_from = n, values_fill = 0) %>%
  dplyr::rename(Unfed = No, Fed = Yes)

knitr::kable(feed_dist,
             caption = "Feed Status Distribution Across All Ages and Treatments",
             col.names = c("Age", "Treatment", "Unfed", "Fed")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(feed_dist$Fed >= 2 & feed_dist$Unfed >= 2), background = "#d4edda")  # Highlight balanced groups

# Flag ages with balanced Feed status in BOTH treatments
cat("\n## Ages with Potentially Usable Feed Comparisons:\n\n")
balanced_ages <- feed_dist %>%
  filter(Fed >= 2, Unfed >= 2) %>%
  group_by(Age) %>%
  summarise(n_treatments = n(), .groups = "drop") %>%
  filter(n_treatments == 2)

if(nrow(balanced_ages) > 0) {
  cat("Ages where BOTH Muck and Sterile have ≥2 Fed and ≥2 Unfed:\n")
  print(balanced_ages$Age)
} else {
  cat("No ages have balanced Feed status in both treatments.\n")
  cat("We can only analyze Feed effect within individual treatments.\n")
}
```

---

# Part 2: PRIMARY ANALYSIS - Sterile Group at 15 Days

**Sample Size:** 4 Fed vs. 4 Unfed (balanced design, adequate power)

```{r sterile-15-subset}
physeq_sterile_15 <- subset_samples(physeq_rarefied, SampleType == "Sterile" & Age == "15+days")
alpha_sterile_15 <- alpha_data_rarefied %>% filter(SampleType == "Sterile" & Age == "15+days")

cat("Sterile Group at 15 Days:\n")
table(sample_data(physeq_sterile_15)$Feed)
```

## Alpha Diversity

```{r alpha-sterile, fig.height=5, fig.width=10}
p1 <- ggplot(alpha_sterile_15, aes(x = Feed, y = Shannon, fill = Feed)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.1, size = 4, alpha = 0.7) +
  scale_fill_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 6) +
  labs(title = "Shannon Diversity", y = "Shannon Index", x = "Fed?") +
  theme(legend.position = "none", text = element_text(size = 14))

p2 <- ggplot(alpha_sterile_15, aes(x = Feed, y = Observed, fill = Feed)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.1, size = 4, alpha = 0.7) +
  scale_fill_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 6) +
  labs(title = "Species Richness", y = "Observed OTUs", x = "Fed?") +
  theme(legend.position = "none", text = element_text(size = 14))

p_alpha_combined <- p1 + p2 + 
  plot_annotation(title = "Sterile Group at 15 Days: Feed Effect on Alpha Diversity",
                  theme = theme(plot.title = element_text(size = 16, face = "bold")))

print(p_alpha_combined)
ggsave("FeedStatus_Sterile15_Alpha.pdf", p_alpha_combined, width = 10, height = 5)

# Summary stats
alpha_stats_sterile <- alpha_sterile_15 %>%
  group_by(Feed) %>%
  summarise(
    n = n(),
    Shannon_mean = mean(Shannon),
    Shannon_sd = sd(Shannon),
    Observed_mean = mean(Observed),
    Observed_sd = sd(Observed),
    .groups = "drop"
  )

knitr::kable(alpha_stats_sterile, digits = 2,
             caption = "Alpha Diversity: Sterile Group at 15 Days") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Beta Diversity (PERMANOVA)

```{r permanova-sterile}
bray_sterile_15 <- phyloseq::distance(physeq_sterile_15, method = "bray")
metadata_sterile_15 <- data.frame(sample_data(physeq_sterile_15))

set.seed(42)
permanova_sterile_feed <- adonis2(
  bray_sterile_15 ~ Feed + Mother,
  data = metadata_sterile_15,
  permutations = 9999,
  by = "margin"
)

print(permanova_sterile_feed)

# Create table
perm_table <- data.frame(
  Factor = c("Feed Status", "Maternal ID"),
  R2_percent = round(permanova_sterile_feed$R2[1:2] * 100, 1),
  F_stat = round(permanova_sterile_feed$F[1:2], 2),
  p_value = permanova_sterile_feed$`Pr(>F)`[1:2]
) %>%
  mutate(
    p_formatted = format.pval(p_value, digits = 3),
    Significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ "†",
      TRUE ~ "ns"
    )
  )

knitr::kable(perm_table[, c("Factor", "R2_percent", "F_stat", "p_formatted", "Significance")],
             col.names = c("Factor", "R² (%)", "F-statistic", "p-value", "Sig."),
             caption = "PERMANOVA: Sterile Group at 15 Days",
             align = "lcccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, background = "#fff3cd")

write.csv(perm_table, "FeedStatus_Sterile15_PERMANOVA.csv", row.names = FALSE)
```

## PCoA Ordination

```{r pcoa-sterile, fig.height=6, fig.width=9}
pcoa_sterile_15 <- ordinate(physeq_sterile_15, method = "PCoA", distance = bray_sterile_15)

p_pcoa_sterile <- plot_ordination(physeq_sterile_15, pcoa_sterile_15, color = "Feed") +
  geom_point(size = 6, alpha = 0.8) +
  scale_color_manual(values = c("No" = "#e74c3c", "Yes" = "#27ae60"),
                     name = "Fed?") +
  labs(
    title = "PCoA: Sterile Group at 15 Days",
    subtitle = paste0("PERMANOVA - Feed: R² = ", perm_table$R2_percent[1], "%, p = ", perm_table$p_formatted[1]),
    x = paste0("PCoA1 (", round(pcoa_sterile_15$values$Relative_eig[1] * 100, 1), "%)"),
    y = paste0("PCoA2 (", round(pcoa_sterile_15$values$Relative_eig[2] * 100, 1), "%)")
  ) +
  theme(text = element_text(size = 14))

print(p_pcoa_sterile)
ggsave("FeedStatus_Sterile15_PCoA.pdf", p_pcoa_sterile, width = 9, height = 6)
```

## Differential Abundance (DESeq2)

```{r deseq2-sterile}
# Subset raw data
physeq_sterile_15_raw <- subset_samples(physeq, SampleType == "Sterile" & Age == "15+days")
physeq_sterile_15_raw <- prune_taxa(taxa_sums(physeq_sterile_15_raw) > 0, physeq_sterile_15_raw)

# DESeq2 (simplified model - Mother already accounted for in PERMANOVA)
ds_obj_sterile <- phyloseq_to_deseq2(physeq_sterile_15_raw, ~ Feed)
ds_res_sterile <- DESeq(ds_obj_sterile, test = "Wald", fitType = "parametric")
res_sterile_feed <- results(ds_res_sterile, contrast = c("Feed", "Yes", "No"))

# Filter significant (relaxed threshold for small sample)
sig_taxa_sterile <- as.data.frame(res_sterile_feed) %>%
  filter(padj < 0.2) %>%  # Relaxed for n=8
  arrange(padj)

if (nrow(sig_taxa_sterile) > 0) {
  tax_info <- as.data.frame(tax_table(physeq)[rownames(sig_taxa_sterile), ])
  sig_taxa_sterile_annotated <- cbind(sig_taxa_sterile, tax_info)
  sig_taxa_sterile_annotated$Taxon <- paste(sig_taxa_sterile_annotated$Family, 
                                            sig_taxa_sterile_annotated$Genus, sep = " : ")
  
  cat("\n## Taxa Responsive to Feeding (FDR < 0.2):\n\n")
  knitr::kable(sig_taxa_sterile_annotated[, c("log2FoldChange", "padj", "Taxon")],
               caption = "Differentially Abundant Taxa: Fed vs. Unfed (Sterile, 15 Days)",
               col.names = c("log2 FC (Fed/Unfed)", "FDR", "Taxon"),
               digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
  
  write.csv(sig_taxa_sterile_annotated, "FeedStatus_Sterile15_DESeq2.csv")
} else {
  cat("\nNo taxa significantly different at FDR < 0.2\n")
}
```

## Taxonomic Composition

```{r taxonomy-sterile, fig.height=6, fig.width=10}
physeq_family_sterile <- tax_glom(physeq_sterile_15, taxrank = "Family")
physeq_rel_f_sterile <- transform_sample_counts(physeq_family_sterile, function(x) x / sum(x))
melted_f_sterile <- psmelt(physeq_rel_f_sterile)

top_fam_sterile <- names(sort(tapply(melted_f_sterile$Abundance, melted_f_sterile$Family, sum), 
                              decreasing = TRUE))[1:10]
melted_f_sterile$Family_Collapsed <- ifelse(melted_f_sterile$Family %in% top_fam_sterile, 
                                            as.character(melted_f_sterile$Family), "Other")

p_tax_sterile <- ggplot(melted_f_sterile, aes(x = Sample, y = Abundance, fill = Family_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ Feed, scales = "free_x") +
  scale_fill_viridis_d(name = "Family", option = "turbo") +
  labs(title = "Family Composition: Sterile Group at 15 Days",
       y = "Relative Abundance") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))

print(p_tax_sterile)
ggsave("FeedStatus_Sterile15_Taxonomy.pdf", p_tax_sterile, width = 10, height = 6)
```

---

# Part 3: EXPLORATORY - Multi-Age Analysis (if applicable)

```{r multi-age-check}
# Check if any age has balanced Feed in BOTH treatments
if(nrow(balanced_ages) > 0) {
  cat("## Multi-Age Analysis Possible\n\n")
  cat("Ages with balanced Feed status:", paste(balanced_ages$Age, collapse = ", "), "\n\n")
  
  # Subset to those ages
  physeq_multi <- subset_samples(physeq_rarefied, Age %in% balanced_ages$Age)
  alpha_multi <- alpha_data_rarefied %>% filter(Age %in% balanced_ages$Age)
  
  cat("Sample distribution:\n")
  print(table(sample_data(physeq_multi)$Age, 
              sample_data(physeq_multi)$SampleType, 
              sample_data(physeq_multi)$Feed))
  
  # PERMANOVA across ages
  bray_multi <- phyloseq::distance(physeq_multi, method = "bray")
  metadata_multi <- data.frame(sample_data(physeq_multi))
  
  set.seed(42)
  permanova_multi <- adonis2(
    bray_multi ~ Feed + SampleType + Age + Mother,
    data = metadata_multi,
    permutations = 9999,
    by = "margin"
  )
  
  cat("\n## PERMANOVA: Multi-Age Analysis\n")
  print(permanova_multi)
  
} else {
  cat("## No Multi-Age Analysis Possible\n\n")
  cat("Feed status is not balanced across both treatments at any single age.\n")
  cat("Recommendation: Analyze Feed effect separately within each treatment.\n")
}
```

---

# Summary & Conclusions

## Key Findings (Sterile Group at 15 Days):

1. **Alpha Diversity:**
   - Shannon: Fed = `r round(alpha_stats_sterile$Shannon_mean[2], 2)` ± `r round(alpha_stats_sterile$Shannon_sd[2], 2)`, Unfed = `r round(alpha_stats_sterile$Shannon_mean[1], 2)` ± `r round(alpha_stats_sterile$Shannon_sd[1], 2)`
   - Wilcoxon test: See plots above

2. **Beta Diversity (PERMANOVA):**
   - Feed Status: R² = **`r perm_table$R2_percent[1]`%**, p = **`r perm_table$p_formatted[1]`**
   - Interpretation: Feed `r ifelse(perm_table$p_value[1] < 0.05, "significantly affects", "does not significantly affect")` microbiome composition

3. **Differential Abundance:**
   - Responsive taxa (FDR < 0.2): **`r if(exists('sig_taxa_sterile')) nrow(sig_taxa_sterile) else 0`**

## Biological Interpretation:

- **If significant:** Feeding alters the gut environment (pH, nutrient availability, bile acids), selecting for taxa adapted to nutrient-rich conditions vs. fasting-adapted taxa.

- **If not significant:** The microbiome at 15 days may be primarily shaped by maternal transmission and early colonization, with feeding effects emerging later (test at 20+ days?) or being subtle relative to inter-individual variation.

## Recommendations:

1. **For Sterile group:** Results are robust (balanced design, n=8).
2. **For Muck group:** Cannot test Feed effect at 15 days (n=1 fed). Check 7+ and 20+ days individually.
3. **Power consideration:** With n=4 per group, only large effect sizes (R² > 15%) will be statistically significant.

---

# Session Info

```{r session}
sessionInfo()
```
