---
title: "Kohler Microbiome Study: Publication Analysis"
author: "Edgardo L. Rosado Ramos"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8)
```
# 1. Setup and Libraries
Load necessary packages for microbiome analysis and visualization.
```{r libraries}
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(DESeq2)
library(indicspecies) # For Indicator Species Analysis
# Set a consistent theme for plots
theme_set(theme_bw(base_size = 12))
```
# 2. Data Loading and Processing
Reading OTU table, taxonomy, and metadata, then creating the phyloseq object.
```{r load-data}
# --- Read OTU Table ---
otu_raw <- read.table("kohler.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group 
otu_mat <- t(otu_mat)
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)
# --- Read Taxonomy ---
tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU
taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
  x <- x[nchar(x) > 0]
  x <- gsub("\\(.*?\\)|\\\"", "", x)
  trimws(x)
})
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)
tax_tab <- tax_table(as.matrix(taxonomy_mat))
# --- Read Metadata ---
Kohler.design <- read.table("Kohlerdesign.txt",
                            header = TRUE,
                            sep = "\t",
                            row.names = 1,
                            stringsAsFactors = TRUE,
                            check.names = FALSE)
# --- Create Phyloseq Object ---
common_samples <- intersect(sample_names(otu_tab), rownames(Kohler.design))
otu_tab_pruned <- prune_samples(common_samples, otu_tab)
sample_data_obj <- sample_data(Kohler.design[common_samples, ])
physeq <- phyloseq(otu_tab_pruned, tax_tab, sample_data_obj)
print(physeq)
```
## Rarefaction and Alpha Diversity Calculation
Filtering low-abundance taxa and rarefying to even depth for diversity analysis.
```{r rarefaction}
# Filter taxa with < 3 counts in < 20% of samples (noise reduction)
physeq_filt <- filter_taxa(physeq, function(x) sum(x > 3) > (0.2 * length(x)), TRUE)
# Rarefy
min_depth <- min(sample_sums(physeq_filt))
set.seed(42)
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)
# Calculate Alpha Diversity
alpha_diversity_rarefied <- estimate_richness(physeq_rarefied, measures = c("Observed", "Shannon"))
sample_metadata_rarefied <- as.data.frame(sample_data(physeq_rarefied))
alpha_data_rarefied <- cbind(alpha_diversity_rarefied, sample_metadata_rarefied)
# Order factors for plotting
age_order <- c("Fetus", "Newborn", "7+days", "15+days", "20+days")
alpha_data_rarefied$Age <- factor(alpha_data_rarefied$Age, levels = age_order)
```
# 3. Sample Size Analysis
Visualizing the distribution of samples across Age and Treatment groups to be transparent about study design limitations.
```{r fig1-sample-dist, fig.height=5, fig.width=8}
# Summarize sample sizes
sample_summary <- alpha_data_rarefied %>%
  group_by(Age, SampleType) %>%
  summarise(n = n(), .groups = "drop")
print(sample_summary)
# Create Heatmap
p_sample_dist <- ggplot(sample_summary, aes(x = Age, y = SampleType, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 6, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#d73027", high = "#1a9850", 
                      name = "Sample\nSize",
                      limits = c(0, 10),
                      breaks = c(0, 2, 4, 6, 8, 10)) +
  labs(title = "Figure 1: Sample Distribution Across Experimental Groups",
       x = "Developmental Stage", y = "Treatment") +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
print(p_sample_dist)
ggsave("Fig1_SampleDistribution.pdf", p_sample_dist, width = 8, height = 5)
```
# 4. Beta Diversity Analysis
Using PERMANOVA to test for effects of SampleType, Age, and Mother.
```{r beta-diversity}
# Calculate Bray-Curtis distance
bray_dist <- phyloseq::distance(physeq_rarefied, method = "bray")
# CRITICAL FIX: Ensure metadata is a standard data frame
# sample_data objects can sometimes cause issues in vegan functions
metadata <- data.frame(sample_data(physeq_rarefied))
# Standard PERMANOVA
set.seed(42)
permanova_standard <- adonis2(
  bray_dist ~ SampleType + Age + Mother,
  data = metadata,
  permutations = 9999,
  by = "margin"
)
print(permanova_standard)
# Check Homogeneity of Dispersion (PERMDISP)
dispersion_sampletype <- betadisper(bray_dist, metadata$SampleType)
dispersion_age <- betadisper(bray_dist, metadata$Age)
dispersion_mother <- betadisper(bray_dist, metadata$Mother)
cat("PERMDISP p-values:\n")
cat("SampleType:", anova(dispersion_sampletype)$`Pr(>F)`[1], "\n")
cat("Age:", anova(dispersion_age)$`Pr(>F)`[1], "\n")
cat("Mother:", anova(dispersion_mother)$`Pr(>F)`[1], "\n")
```
## Effect Size Summary
Calculating and tabulating effect sizes (R²) to emphasize biological relevance over just p-values.
```{r effect-sizes}
effect_size_table <- data.frame(
  Factor = c("Mother", "SampleType", "Age"),
  R2 = c(
    permanova_standard$R2[which(rownames(permanova_standard) == "Mother")],
    permanova_standard$R2[which(rownames(permanova_standard) == "SampleType")],
    permanova_standard$R2[which(rownames(permanova_standard) == "Age")]
  ),
  F_stat = c(
    permanova_standard$F[which(rownames(permanova_standard) == "Mother")],
    permanova_standard$F[which(rownames(permanova_standard) == "SampleType")],
    permanova_standard$F[which(rownames(permanova_standard) == "Age")]
  ),
  p_value = c(
    permanova_standard$`Pr(>F)`[which(rownames(permanova_standard) == "Mother")],
    permanova_standard$`Pr(>F)`[which(rownames(permanova_standard) == "SampleType")],
    permanova_standard$`Pr(>F)`[which(rownames(permanova_standard) == "Age")]
  )
) %>%
  mutate(
    R2_percent = round(R2 * 100, 1),
    Effect_Size = case_when(
      R2 < 0.01 ~ "Negligible",
      R2 < 0.06 ~ "Small",
      R2 < 0.14 ~ "Medium",
      TRUE ~ "Large"
    ),
    Significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )
knitr::kable(effect_size_table, caption = "Table 1: Effect Size Summary")
write.csv(effect_size_table, "Table1_EffectSizes.csv", row.names = FALSE)
```
# 5. Figure 2: Maternal Effects
Visualizing the strong maternal effect (R² ~ 20%) using PCoA, dispersion, and alpha diversity.
```{r fig2-maternal, fig.height=10, fig.width=12}
# Panel A: PCoA colored by Mother
pcoa_res <- ordinate(physeq_rarefied, method = "PCoA", distance = bray_dist)
pcoa_mother <- plot_ordination(physeq_rarefied, pcoa_res, 
                               color = "Mother", shape = "SampleType") +
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = Mother), linetype = 2, linewidth = 0.5) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "A) Community Composition by Maternal Clutch",
       subtitle = paste0("PERMANOVA: R² = ", 
                        round(effect_size_table$R2[1], 3), 
                        ", p = ", 
                        format.pval(effect_size_table$p_value[1], digits = 3)),
       x = paste0("PCoA1 (", round(pcoa_res$values$Relative_eig[1] * 100, 1), "%)"),
       y = paste0("PCoA2 (", round(pcoa_res$values$Relative_eig[2] * 100, 1), "%)"))
# Panel B: Dispersion within clutches
dispersion_df <- data.frame(
  Mother = metadata$Mother,
  Distance = dispersion_mother$distances
)
p_dispersion <- ggplot(dispersion_df, aes(x = Mother, y = Distance, fill = Mother)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "B) Within-Clutch Microbiome Variation",
       y = "Distance to Clutch Centroid (Bray-Curtis)",
       x = "Maternal ID") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
# Panel C: Shannon diversity by Mother
p_alpha_mother <- ggplot(alpha_data_rarefied, 
                         aes(x = Mother, y = Shannon, fill = Mother)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set1") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, 
               fill = "white", color = "black") +
  labs(title = "C) Alpha Diversity by Maternal Clutch",
       y = "Shannon Diversity Index",
       x = "Maternal ID") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
# Combine panels
fig2 <- pcoa_mother / (p_dispersion | p_alpha_mother) +
  plot_annotation(
    title = "Figure 2: Maternal Identity Shapes Offspring Gut Microbiomes",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
print(fig2)
ggsave("Fig2_MaternalEffects.pdf", fig2, width = 12, height = 10)
```
# 6. Figure 3: Environmental Effects
Focusing on the well-powered age groups (7+days and 15+days) to show environmental effects.
```{r fig3-env, fig.height=6, fig.width=14}
# Focus on well-powered age groups
physeq_subset <- subset_samples(physeq_rarefied, 
                                Age %in% c("7+days", "15+days"))
bray_subset <- phyloseq::distance(physeq_subset, method = "bray")
pcoa_subset <- ordinate(physeq_subset, method = "PCoA", distance = bray_subset)
# Panel A: PCoA by SampleType
pcoa_env <- plot_ordination(physeq_subset, pcoa_subset, 
                            color = "SampleType", shape = "Age") +
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = SampleType), linetype = 2, linewidth = 0.8) +
  scale_color_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  labs(title = "A) Community Composition by Environmental Exposure",
       subtitle = paste0("PERMANOVA: R² = ", 
                        round(effect_size_table$R2[2], 3), 
                        ", p = ", 
                        format.pval(effect_size_table$p_value[2], digits = 3),
                        " (7+days and 15+days only)"),
       x = paste0("PCoA1 (", round(pcoa_subset$values$Relative_eig[1] * 100, 1), "%)"),
       y = paste0("PCoA2 (", round(pcoa_subset$values$Relative_eig[2] * 100, 1), "%)"))
# Panel B: Alpha diversity by SampleType
alpha_subset <- alpha_data_rarefied %>%
  filter(Age %in% c("7+days", "15+days"))
p_alpha_env <- ggplot(alpha_subset, aes(x = SampleType, y = Shannon, 
                                        fill = SampleType)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  labs(title = "B) Shannon Diversity by Treatment",
       subtitle = "7+days and 15+days only",
       y = "Shannon Diversity Index",
       x = "Treatment") +
  theme(legend.position = "none")
fig3 <- pcoa_env | p_alpha_env
print(fig3)
ggsave("Fig3_EnvironmentalEffects.pdf", fig3, width = 14, height = 6)
```
# 8. Taxonomic Composition
Since no specific taxa were significantly different by strict statistical thresholds (likely due to sample size), we will visualize the community structure in more detail to identify qualitative patterns.
## A) Phylum Level Composition
```{r fig4a-phylum, fig.height=6, fig.width=12}
# Agglomerate to Phylum
physeq_phylum <- tax_glom(physeq_rarefied, taxrank = "Phylum")
physeq_rel_p <- transform_sample_counts(physeq_phylum, function(x) x / sum(x))
melted_phy_p <- psmelt(physeq_rel_p)
# Top 10 Phyla
top_phyla <- names(sort(tapply(melted_phy_p$Abundance, melted_phy_p$Phylum, sum), decreasing = TRUE))[1:10]
melted_phy_p$Phylum_Collapsed <- ifelse(melted_phy_p$Phylum %in% top_phyla, as.character(melted_phy_p$Phylum), "Other")
p_phylum <- ggplot(melted_phy_p, aes(x = Sample, y = Abundance, fill = Phylum_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ Age, scales = "free_x", space = "free") +
  scale_fill_brewer(palette = "Paired", name = "Phylum") +
  labs(title = "Figure 4A: Phylum Composition", y = "Relative Abundance") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
print(p_phylum)
```
## B) Family Level Composition
```{r fig4b-family, fig.height=6, fig.width=12}
# Agglomerate to Family
physeq_family <- tax_glom(physeq_rarefied, taxrank = "Family")
physeq_rel_f <- transform_sample_counts(physeq_family, function(x) x / sum(x))
melted_phy_f <- psmelt(physeq_rel_f)
# Top 15 Families
top_families <- names(sort(tapply(melted_phy_f$Abundance, melted_phy_f$Family, sum), decreasing = TRUE))[1:15]
melted_phy_f$Family_Collapsed <- ifelse(melted_phy_f$Family %in% top_families, as.character(melted_phy_f$Family), "Other")
p_family <- ggplot(melted_phy_f, aes(x = Sample, y = Abundance, fill = Family_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ Age, scales = "free_x", space = "free") +
  scale_fill_viridis_d(name = "Family", option = "turbo") + # Using turbo for more distinct colors
  labs(title = "Figure 4B: Family Composition", y = "Relative Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8))
print(p_family)
ggsave("Fig4_Taxonomy_Family.pdf", p_family, width = 12, height = 6)
```
## C) Heatmap of Top Genera
```{r fig5-heatmap, fig.height=8, fig.width=10}
# Filter to top 20 Genera
physeq_genus <- tax_glom(physeq_rarefied, taxrank = "Genus")
top20_genera <- names(sort(taxa_sums(physeq_genus), decreasing = TRUE))[1:20]
physeq_top20 <- prune_taxa(top20_genera, physeq_genus)
# Plot Heatmap
p_heatmap <- plot_heatmap(physeq_top20, 
                          method = "NMDS", 
                          distance = "bray", 
                          sample.label = "SampleType", 
                          taxa.label = "Genus") +
  facet_grid(~ Age, scales = "free_x", space = "free") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "white") +
  labs(title = "Figure 5: Heatmap of Top 20 Genera") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(p_heatmap)
ggsave("Fig5_Heatmap.pdf", p_heatmap, width = 10, height = 8)
```
# 9. Differential Abundance (DESeq2)
Identifying specific taxa that differ between Muck and Sterile treatments (controlling for Age).
```{r deseq2}
# Convert phyloseq to DESeq2 object
# CRITICAL FIX: Subset to ages where both treatments exist to avoid "not full rank" error
# Fetus and Newborn only have one SampleType, which causes confounding.
physeq_deseq <- subset_samples(physeq, Age %in% c("7+days", "15+days", "20+days"))
# Remove taxa that are not present in this subset
physeq_deseq <- prune_taxa(taxa_sums(physeq_deseq) > 0, physeq_deseq)
ds_obj <- phyloseq_to_deseq2(physeq_deseq, ~ Age + SampleType)
# Run DESeq2
ds_res <- DESeq(ds_obj, test = "Wald", fitType = "parametric")
# Get results for Muck vs Sterile
res <- results(ds_res, contrast = c("SampleType", "Muck", "Sterile"))
# Filter for significance
sig_taxa <- as.data.frame(res) %>%
  filter(padj < 0.05) %>%
  arrange(desc(log2FoldChange))
# Add taxonomy
if (nrow(sig_taxa) > 0) {
  tax_info <- as.data.frame(tax_table(physeq)[rownames(sig_taxa), ])
  sig_taxa_annotated <- cbind(sig_taxa, tax_info)
  
  print("Differentially Abundant Taxa (Muck vs. Sterile):")
  knitr::kable(sig_taxa_annotated[, c("log2FoldChange", "padj", "Phylum", "Family", "Genus")], 
               caption = "Table 3: Significant Differentially Abundant Taxa")
  write.csv(sig_taxa_annotated, "Table3_DifferentialAbundance.csv")
} else {
  print("No taxa found to be significantly different at FDR < 0.05")
}
```
# 10. Indicator Species Analysis
An alternative to DESeq2 that looks for taxa that are "indicators" of a specific group (high fidelity and specificity). This is often more sensitive for presence/absence patterns.
```{r indicator-species}
# Subset to relevant ages (7+days, 15+days, 20+days)
physeq_indic <- subset_samples(physeq_rarefied, Age %in% c("7+days", "15+days", "20+days"))
physeq_indic <- prune_taxa(taxa_sums(physeq_indic) > 0, physeq_indic)
# Extract OTU table (species as columns)
otu_indic <- as(otu_table(physeq_indic), "matrix")
if(taxa_are_rows(physeq_indic)) { otu_indic <- t(otu_indic) }
# Extract grouping vector
groups_indic <- sample_data(physeq_indic)$SampleType
# Run Indicator Species Analysis
indval <- multipatt(otu_indic, groups_indic, control = how(nperm=999))
# Summary of significant indicators
summary(indval, alpha = 0.05)
# Extract and annotate significant results
if (!is.null(indval$sign)) {
  sig_ind <- indval$sign %>%
    filter(p.value < 0.05)
  
  if (nrow(sig_ind) > 0) {
    # Add taxonomy
    tax_ind <- as.data.frame(tax_table(physeq_indic)[rownames(sig_ind), ])
    results_ind <- cbind(sig_ind, tax_ind)
    
    print("Significant Indicator Species:")
    knitr::kable(results_ind[, c("stat", "p.value", "Phylum", "Family", "Genus")], 
                 caption = "Table 4: Indicator Species Analysis")
    write.csv(results_ind, "Table4_IndicatorSpecies.csv")
  } else {
    print("No significant indicator species found.")
  }
}
```
# 11. Core Microbiome Analysis
Identifying the "Core" microbiome: taxa present in at least 50% of samples within a treatment group. This defines the stable resident community.
```{r core-microbiome, fig.height=8, fig.width=10}
# Function to get core taxa names
get_core_taxa <- function(physeq_obj, detection_threshold = 0, prevalence_threshold = 0.5) {
  # Transform to relative abundance
  physeq_rel <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  
  # Filter
  core_taxa <- filter_taxa(physeq_rel, function(x) sum(x > detection_threshold) > (prevalence_threshold * length(x)), TRUE)
  return(taxa_names(core_taxa))
}
# Get core for Muck
physeq_muck <- subset_samples(physeq_rarefied, SampleType == "Muck" & Age %in% c("7+days", "15+days", "20+days"))
core_muck <- get_core_taxa(physeq_muck)
# Get core for Sterile
physeq_sterile <- subset_samples(physeq_rarefied, SampleType == "Sterile" & Age %in% c("7+days", "15+days", "20+days"))
core_sterile <- get_core_taxa(physeq_sterile)
# Find shared and unique core taxa
shared_core <- intersect(core_muck, core_sterile)
unique_muck <- setdiff(core_muck, core_sterile)
unique_sterile <- setdiff(core_sterile, core_muck)
cat("Core Taxa Summary (Prevalence > 50%):\n")
cat("Shared Core:", length(shared_core), "taxa\n")
cat("Unique to Muck Core:", length(unique_muck), "taxa\n")
cat("Unique to Sterile Core:", length(unique_sterile), "taxa\n")
# Visualize Core Taxa (Heatmap of these specific taxa)
all_core <- c(shared_core, unique_muck, unique_sterile)
if(length(all_core) > 0) {
  physeq_core <- prune_taxa(all_core, physeq_rarefied)
  
  p_core <- plot_heatmap(physeq_core, 
                         method = "NMDS", 
                         distance = "bray", 
                         sample.label = "SampleType", 
                         taxa.label = "Genus") +
    facet_grid(~ SampleType, scales = "free_x", space = "free") +
    scale_fill_gradient(low = "white", high = "darkgreen", na.value = "white") +
    labs(title = "Figure 6: Core Microbiome Heatmap (Prevalence > 50%)")
  
  print(p_core)
  ggsave("Fig6_CoreMicrobiome.pdf", p_core, width = 10, height = 8)
  
  # Save list of core taxa
  core_tax_table <- as.data.frame(tax_table(physeq_core))
  core_tax_table$Core_Status <- case_when(
    rownames(core_tax_table) %in% shared_core ~ "Shared",
    rownames(core_tax_table) %in% unique_muck ~ "Muck_Only",
    rownames(core_tax_table) %in% unique_sterile ~ "Sterile_Only"
  )
  write.csv(core_tax_table, "Table5_CoreTaxa.csv")
}
```
# 12. Summary Statistics
```{r summary-stats}
summary_stats <- alpha_data_rarefied %>%
  group_by(SampleType, Age) %>%
  summarise(
    n = n(),
    Shannon_mean = mean(Shannon, na.rm = TRUE),
    Shannon_sd = sd(Shannon, na.rm = TRUE),
    Shannon_se = sd(Shannon, na.rm = TRUE) / sqrt(n()),
    Observed_mean = mean(Observed, na.rm = TRUE),
    Observed_sd = sd(Observed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Shannon_summary = paste0(round(Shannon_mean, 2), " ± ", round(Shannon_se, 2)),
    Observed_summary = paste0(round(Observed_mean, 1), " ± ", round(Observed_sd, 1))
  )
knitr::kable(summary_stats, caption = "Table 2: Summary Statistics")
write.csv(summary_stats, "Table2_SummaryStats.csv", row.names = FALSE)
```
# 11. Session Info
```{r session-info}
sessionInfo()
```