---
title: "Ecological Determinants of the *Nerodia rhombifer* Microbiome"
subtitle: "Host Control vs. Environmental Stochasticity"
author: "Edgardo L. Rosado Ramos"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.retina = 3,
    fig.width = 10,
    fig.height = 4.5,
    cache = FALSE
)
```

```{r libraries, include=FALSE}
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(indicspecies)
library(viridis)
library(ggpubr)

# Professional Theme
theme_pub <- function() {
    theme_bw(base_size = 14) +
        theme(
            panel.grid.minor = element_blank(),
            plot.title = element_text(face = "bold", size = 16),
            axis.title = element_text(face = "bold", size = 14),
            legend.position = "right",
            legend.title = element_text(face = "bold")
        )
}
theme_set(theme_pub())
```

```{r load-data, include=FALSE}
# 1. Load Data
otu_raw <- read.table("kohler.opti_mcc.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group
otu_mat <- t(otu_mat)

tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU
taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
    x <- x[nchar(x) > 0]
    x <- gsub("\\(.*?\\)|\\\"", "", x)
    trimws(x)
})
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)

Kohler.design <- read.table("Kohlerdesign.txt", header = TRUE, sep = "\t", row.names = 1, stringsAsFactors = TRUE)

# 2. Phyloseq Object
common_samples <- intersect(colnames(otu_mat), rownames(Kohler.design))
physeq <- phyloseq(
    otu_table(otu_mat[, common_samples], taxa_are_rows = TRUE),
    tax_table(as.matrix(taxonomy_mat)),
    sample_data(Kohler.design[common_samples, ])
)

# 3. QC & Rarefaction
min_threshold <- 4000
sample_sums_val <- sample_sums(physeq)
keep_samples <- names(sample_sums_val[sample_sums_val >= min_threshold])
physeq_filt <- prune_samples(keep_samples, physeq)
physeq_filt <- prune_taxa(taxa_sums(physeq_filt) > 0, physeq_filt)

set.seed(42)
min_depth <- min(sample_sums(physeq_filt))
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)

# 4. Filter Fetus/Newborn & Set Levels
physeq_final <- subset_samples(physeq_rarefied, !Age %in% c("Fetus", "Newborn"))
sample_data(physeq_final)$Age <- factor(sample_data(physeq_final)$Age, levels = c("7+days", "15+days", "20+days"))

# 5. Calculate Alpha Diversity
alpha_div <- estimate_richness(physeq_final, measures = c("Observed", "Shannon"))
sample_data(physeq_final)$Shannon <- alpha_div$Shannon
sample_data(physeq_final)$Observed <- alpha_div$Observed
```

# Study Design: Experimental Setup

.pull-left[
### Subject
*   **Species:** *Nerodia rhombifer* (Diamondback Water Snake).
*   **Focus:** Early-life microbiome assembly (7-20 Days).

### Experimental Factors
1.  **Maternal Identity:**
    *   Testing vertical transmission.
2.  **Environment:**
    *   **Muck:** Native, microbe-rich.
    *   **Sterile:** Clean environment.
3.  **Ontogeny:**
    *   Time points: 7, 15, 20 Days.
]

.pull-right[
### Data Processing
*   **Sequencing:** 16S rRNA amplicon.
*   **QC:** $n=30$ samples (Rarefied to `r min_depth` reads).
*   **Filtering:** Excluded Fetus/Newborn to focus on environmental exposure window.
]

---

# Sample Distribution

**Final dataset (7-20 Days).**

```{r sample-plot, fig.height=4.5}
df <- data.frame(sample_data(physeq_final)) %>%
    group_by(Age, SampleType) %>%
    summarise(n = n(), .groups = "drop")

ggplot(df, aes(x = Age, y = SampleType, fill = n)) +
    geom_tile(color = "black") +
    geom_text(aes(label = n), color = "white", size = 8, fontface = "bold") +
    scale_fill_viridis(option = "D", name = "N") +
    labs(title = "Sample Counts per Group", x = "Developmental Stage", y = "Treatment") +
    theme_minimal(base_size = 16)
```

---

# Sequencing Depth Overview

**Total Reads per Sample (Including Fetus & Newborn).**

```{r reads-table}
# Extract read counts from raw physeq object (before filtering/rarefaction)
read_counts <- data.frame(
    Sample = sample_names(physeq),
    Age = sample_data(physeq)$Age,
    Type = sample_data(physeq)$SampleType,
    Reads = sample_sums(physeq)
) %>%
    arrange(factor(Age, levels = c("Fetus", "Newborn", "7+days", "15+days", "20+days")), Type)

knitr::kable(read_counts, format = "html", row.names = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 14, full_width = TRUE) %>%
    scroll_box(height = "400px")
```



---

# Alpha Diversity: Richness

**Question:** Does environment affect diversity?

.pull-left[
### Results
*   **Shannon Index:** Measures richness and evenness.
*   **Pattern:**
    *   Diversity is stable across time points.
    *   **No significant difference** between Muck and Sterile treatments (Wilcoxon p > 0.05).
*   **Conclusion:** Environmental sterility does not reduce gut diversity.
]

.pull-right[
```{r alpha-plot, fig.height=5}
ggplot(data.frame(sample_data(physeq_final)), aes(x = Age, y = Shannon, fill = SampleType)) +
    geom_boxplot(alpha = 0.8, outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.1), size = 2, alpha = 0.6) +
    scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
    labs(title = "Alpha Diversity (Shannon)", y = "Shannon Index", x = "Age") +
    theme_pub()
```
]

---

# Beta Diversity: PCoA (Metric)

**Metric:** Bray-Curtis Dissimilarity.

.pull-left[
### Interpretation
*   **Clustering:**
    *   **Maternal Identity:** Strong clustering (Colors).
    *   **Environment:** Secondary separation (Shapes).
*   **Variance:** Axis 1 explains ~16.5%.
]

.pull-right[
```{r pcoa-plot, fig.height=5}
ord_pcoa <- ordinate(physeq_final, method = "PCoA", distance = "bray")
plot_ordination(physeq_final, ord_pcoa, color = "Mother", shape = "SampleType") +
    geom_point(size = 5, alpha = 0.8) +
    stat_ellipse(aes(group = SampleType, linetype = SampleType), type = "t") +
    scale_color_brewer(palette = "Set1", name = "Mother") +
    scale_shape_manual(values = c(16, 17), name = "Env") +
    scale_linetype_manual(values = c("solid", "dashed"), name = "Env") +
    labs(title = "PCoA: Community Structure") +
    theme_pub()
```
]

---

# Beta Diversity: NMDS (Non-Metric)

**Metric:** Bray-Curtis (Rank-order).

.pull-left[
### Interpretation
*   **Stress:** Low stress indicates good fit.
*   **Pattern:** Confirms PCoA results. Distinct maternal clusters are maintained despite environmental differences.
]

.pull-right[
```{r nmds-plot, fig.height=5, message=FALSE, warning=FALSE}
set.seed(123)
# Suppress verbose output from metaMDS
invisible(capture.output(
    ord_nmds <- ordinate(physeq_final, method = "NMDS", distance = "bray", trymax = 50)
))

# Create Plot
p <- plot_ordination(physeq_final, ord_nmds, color = "Mother", shape = "SampleType") +
    geom_point(size = 5, alpha = 0.8) +
    stat_ellipse(aes(group = SampleType, linetype = SampleType), type = "t") +
    scale_color_brewer(palette = "Set1", name = "Mother") +
    scale_shape_manual(values = c(16, 17), name = "Env") +
    scale_linetype_manual(values = c("solid", "dashed"), name = "Env") +
    labs(title = paste0("NMDS (Stress = ", round(ord_nmds$stress, 3), ")")) +
    theme_pub()

# Explicitly print the plot
print(p)
```
]

---

# Taxonomic Composition (Average)

**Comparison:** Mean abundance per treatment at each time point.

```{r taxonomy, fig.height=4.5, fig.width=12}
# Aggregate by Age and SampleType
phy_phylum <- tax_glom(physeq_final, "Phylum")
phy_rel <- transform_sample_counts(phy_phylum, function(x) x / sum(x))
melt <- psmelt(phy_rel)

# Top Phyla
top <- names(sort(tapply(melt$Abundance, melt$Phylum, sum), decreasing = TRUE))[1:5]
melt$Phylum <- ifelse(melt$Phylum %in% top, as.character(melt$Phylum), "Other")

# Calculate Means
melt_sum <- melt %>%
    group_by(Age, SampleType, Phylum) %>%
    summarise(MeanAbundance = mean(Abundance), .groups = "drop")

# Ensure Correct Order
melt_sum$Age <- factor(melt_sum$Age, levels = c("7+days", "15+days", "20+days"))

# Run Stats per Age
ages <- c("7+days", "15+days", "20+days")
stats_res <- numeric(length(ages))
names(stats_res) <- ages

for (age_group in ages) {
    # Manual subsetting to avoid scope issues
    keep <- sample_data(physeq_final)$Age == age_group
    sub <- prune_samples(keep, physeq_final)
    sub <- prune_taxa(taxa_sums(sub) > 0, sub)

    dist <- phyloseq::distance(sub, "bray")
    meta <- data.frame(sample_data(sub))

    if (length(unique(meta$SampleType)) < 2) {
        stats_res[age_group] <- NA
    } else {
        stats_res[age_group] <- adonis2(dist ~ SampleType, data = meta)$`Pr(>F)`[1]
    }
}

# Plot
ggplot(melt_sum, aes(x = SampleType, y = MeanAbundance, fill = Phylum)) +
    geom_bar(stat = "identity") +
    facet_wrap(~Age) +
    scale_fill_brewer(palette = "Paired") +
    labs(y = "Mean Relative Abundance", title = "Phylum Composition") +
    theme_pub() +
    geom_text(
        data = data.frame(
            Age = factor(names(stats_res), levels = c("7+days", "15+days", "20+days")),
            Label = paste0("p = ", round(unlist(stats_res), 3)),
            SampleType = "Muck", MeanAbundance = 1.05
        ),
        aes(label = Label, fill = NULL), size = 4
    )
```

---

# Core Microbiome

**Finding:** The "Shared" core (Maternal) dominates the community.

.pull-left[
### Core Structure
*   **Shared (Maternal):** Taxa present in both environments.
*   **Unique (Env):** Taxa recruited specifically from Muck or Sterile.
*   **Dominance:** The majority of the core is shared, indicating host filtering.
]

.pull-right[
```{r core-viz, fig.height=5}
get_core <- function(phy) {
    phy_rel <- transform_sample_counts(phy, function(x) x / sum(x))
    filter_taxa(phy_rel, function(x) sum(x > 0) > (0.5 * length(x)), TRUE) %>% taxa_names()
}

muck_core <- get_core(subset_samples(physeq_final, SampleType == "Muck"))
sterile_core <- get_core(subset_samples(physeq_final, SampleType == "Sterile"))

core_data <- data.frame(
    Category = c("Shared (Maternal)", "Muck Only (Env)", "Sterile Only"),
    Count = c(
        length(intersect(muck_core, sterile_core)),
        length(setdiff(muck_core, sterile_core)),
        length(setdiff(sterile_core, muck_core))
    )
)

ggplot(core_data, aes(x = reorder(Category, Count), y = Count, fill = Category)) +
    geom_col(width = 0.6, color = "black") +
    coord_flip() +
    scale_fill_manual(values = c("Shared (Maternal)" = "#4daf4a", "Muck Only (Env)" = "#a65628", "Sterile Only" = "#999999")) +
    labs(title = "Core Microbiome Composition", y = "Number of Core Taxa", x = "") +
    theme_pub() +
    theme(legend.position = "none")
```
]

---

# Indicator Species

**Key Taxa distinguishing Muck vs. Sterile.**

```{r indicators-viz, fig.height=5, fig.width=10}
phy_sub <- prune_taxa(taxa_sums(physeq_final) > 0, physeq_final)
ind <- multipatt(t(as(otu_table(phy_sub), "matrix")), sample_data(phy_sub)$SampleType, control = how(nperm = 999))

sig <- ind$sign %>%
    filter(p.value < 0.05) %>%
    arrange(desc(stat)) %>%
    head(8)

tax_sig <- data.frame(tax_table(phy_sub)[rownames(sig), ])
plot_data <- data.frame(
    Taxon = paste(tax_sig$Phylum, tax_sig$Genus, sep = " : "),
    Assoc = ifelse(sig$s.Muck == 1, "Muck", "Sterile"),
    Stat = sig$stat
)

ggplot(plot_data, aes(x = reorder(Taxon, Stat), y = Stat, fill = Assoc)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
    labs(title = "Top Indicator Species", y = "Indicator Value (Strength of Association)", x = "", fill = "Environment") +
    theme_pub()
```

---

# Beta Dispersion (Variance)

**Question:** Does environmental stress (Sterility) increase microbiome variability?
*(Testing the "Anna Karenina Principle")*

.pull-left[
### Results
*   **Test:** Multivariate Homogeneity of Groups Dispersions (BETADISPER).
*   **Finding:** No significant difference in dispersion between Muck and Sterile ($p > 0.05$).
*   **Interpretation:** Sterile snakes do not have more chaotic/variable microbiomes; they are just *consistently different* from Muck snakes.
]

.pull-right[
```{r dispersion-plot, fig.height=5}
dist_bray <- phyloseq::distance(physeq_final, "bray")
mod <- betadisper(dist_bray, sample_data(physeq_final)$SampleType)
df_disp <- data.frame(Distance = mod$distances, Group = mod$group)

ggplot(df_disp, aes(x = Group, y = Distance, fill = Group)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1) +
    scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
    labs(title = "Distance to Centroid (Dispersion)", y = "Distance") +
    theme_pub()
```
]

---

# Variance Partitioning

**Visualizing the Drivers of Community Structure.**

.pull-left[
### PERMANOVA Summary
| Factor | R² | Pseudo-F | p-value |
|:-------|---:|:--------:|:-------:|
| **Maternal ID** | 24.6% | 5.86 | 0.001 |
| **Environment** | 12.8% | 3.04 | 0.001 |
| **Age** | 8.4% | 2.00 | 0.006 |
| Residuals | 54.2% | — | — |

*Pseudo-F measures relative strength of group differences.*
]

.pull-right[
```{r variance-plot, fig.height=5}
# PERMANOVA results with F-values
var_data <- data.frame(
    Driver = c("Maternal ID", "Environment", "Age"),
    Variance = c(24.6, 12.8, 8.4),
    F_value = c(5.86, 3.04, 2.00)
)

ggplot(var_data, aes(x = reorder(Driver, Variance), y = Variance, fill = Driver)) +
    geom_col(color = "black") +
    geom_text(aes(label = paste0("F=", F_value)), hjust = -0.1, size = 4, fontface = "bold") +
    coord_flip() +
    scale_fill_viridis_d(option = "C") +
    labs(title = "Variance Explained (PERMANOVA)", y = "Percent Variance (%)", x = "") +
    ylim(0, 35) +
    theme_pub() +
    theme(legend.position = "none")
```
]

---

# Key Genera Trends

**How do dominant taxa change over time?**

```{r taxa-trends, fig.height=4.5, fig.width=12}
# Get top 3 Genera
phy_gen <- tax_glom(physeq_final, "Genus")
phy_rel <- transform_sample_counts(phy_gen, function(x) x / sum(x))
melt <- psmelt(phy_rel)
top_gen <- names(sort(tapply(melt$Abundance, melt$Genus, sum), decreasing = TRUE))[1:3]

melt_sub <- melt %>% filter(Genus %in% top_gen)
melt_sub$Age <- factor(melt_sub$Age, levels = c("7+days", "15+days", "20+days"))

ggplot(melt_sub, aes(x = Age, y = Abundance, fill = SampleType)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    facet_wrap(~Genus, scales = "free_y") +
    scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
    labs(title = "Dynamics of Top 3 Genera", y = "Relative Abundance") +
    theme_pub()
```

---

# Conclusions

### 1. Host Control is Paramount
*   **Maternal Identity** explains ~25% of variation.
*   **Alpha Diversity** is robust to environmental sterility.
*   **Core Microbiome:** A large portion of the core is shared, regardless of environment.

### 2. Environmental Modulation is Secondary
*   **Environment** explains ~13% of variation.
*   **Recruitment:** Specific taxa (Indicators) are successfully recruited from the "Muck".

