---
title: "Microbiome Analysis Results"
subtitle: "Maternal vs. Environmental Effects on Gut Development"
author: "Edgardo L. Rosado Ramos"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.retina = 3,
  fig.width = 10,
  fig.height = 6,
  cache = TRUE
)
```
```{r libraries, include=FALSE}
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(kableExtra)
library(DESeq2)
library(indicspecies)
theme_set(theme_bw(base_size = 16))
```
```{r load-data, cache=TRUE, include=FALSE}
# Load OTU Table
otu_raw <- read.table("kohler.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group 
otu_mat <- t(otu_mat)
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)
# Load Taxonomy
tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU
taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
  x <- x[nchar(x) > 0]
  x <- gsub("\\(.*?\\)|\\\"", "", x)
  trimws(x)
})
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)
tax_tab <- tax_table(as.matrix(taxonomy_mat))
# Load Metadata
Kohler.design <- read.table("Kohlerdesign.txt",
                            header = TRUE, sep = "\t", row.names = 1,
                            stringsAsFactors = TRUE, check.names = FALSE)
# Create Phyloseq
common_samples <- intersect(sample_names(otu_tab), rownames(Kohler.design))
otu_tab_pruned <- prune_samples(common_samples, otu_tab)
sample_data_obj <- sample_data(Kohler.design[common_samples, ])
physeq <- phyloseq(otu_tab_pruned, tax_tab, sample_data_obj)
# Rarefy
physeq_filt <- filter_taxa(physeq, function(x) sum(x > 3) > (0.2 * length(x)), TRUE)
min_depth <- min(sample_sums(physeq_filt))
set.seed(42)
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)
# Alpha Diversity
alpha_diversity_rarefied <- estimate_richness(physeq_rarefied, measures = c("Observed", "Shannon"))
sample_metadata_rarefied <- as.data.frame(sample_data(physeq_rarefied))
alpha_data_rarefied <- cbind(alpha_diversity_rarefied, sample_metadata_rarefied)
age_order <- c("Fetus", "Newborn", "7+days", "15+days", "20+days")
alpha_data_rarefied$Age <- factor(alpha_data_rarefied$Age, levels = age_order)
```
class: center, middle
# Study Overview
---
## Research Question
.pull-left[
**Objective:**
Investigate the effects of:
- **Maternal Identity** (vertical transmission)
- **Environmental Exposure** (Muck vs. Sterile)
on the developing gut microbiome
**Sample Size:** n = 32 samples across 5 developmental stages
]
.pull-right[
**Key Findings:**
✓ Strong maternal effect (R² ~ 20%)
✓ Significant environmental effect in later stages
✓ Identified specific "Core" and "Indicator" taxa
]
---
## Sample Distribution
```{r fig1, fig.height=5, fig.width=10}
sample_summary <- alpha_data_rarefied %>%
  group_by(Age, SampleType) %>%
  summarise(n = n(), .groups = "drop")
ggplot(sample_summary, aes(x = Age, y = SampleType, fill = n)) +
  geom_tile(color = "white", size = 1.5) +
  geom_text(aes(label = n), size = 8, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#d73027", high = "#1a9850", limits = c(0, 10)) +
  labs(title = "Samples per Experimental Group", x = "Developmental Stage", y = "Treatment") +
  theme_minimal(base_size = 18) +
  theme(panel.grid = element_blank())
```
.footnote[
**Statistical Methods:**  
- Rarefied to `r min_depth` reads/sample  
- Taxa filtered (≥3 reads in ≥20% of samples)  
- Strategy: Focus on well-powered groups (7+, 15+, 20+ days)  
- PERMANOVA with 9,999 permutations
]
---
## Result 1: Maternal Clustering
```{r fig2, fig.height=6, fig.width=11}
bray_dist <- phyloseq::distance(physeq_rarefied, method = "bray")
metadata <- data.frame(sample_data(physeq_rarefied))
pcoa_res <- ordinate(physeq_rarefied, method = "PCoA", distance = bray_dist)
plot_ordination(physeq_rarefied, pcoa_res, color = "Mother") +
  geom_point(size = 6, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Community Composition Clusters by Maternal ID",
    subtitle = "PERMANOVA: R² = 0.20, p < 0.001",
    x = paste0("PCoA1 (", round(pcoa_res$values$Relative_eig[1] * 100, 1), "%)"),
    y = paste0("PCoA2 (", round(pcoa_res$values$Relative_eig[2] * 100, 1), "%)")
  ) +
  theme(legend.position = "right", legend.text = element_text(size = 14))
```
.footnote[**Interpretation:** Strong vertical transmission or shared early environment]
---
## Statistical Drivers of Microbiome Variation
```{r permanova-table}
# Run PERMANOVA
set.seed(42)
permanova_full <- adonis2(
  bray_dist ~ Mother + SampleType + Age,
  data = metadata,
  permutations = 9999,
  by = "margin"
)
# Create effect size table
effect_table <- data.frame(
  Factor = c("Maternal ID", "Treatment (Muck/Sterile)", "Developmental Age"),
  R2 = c(
    permanova_full$R2[which(rownames(permanova_full) == "Mother")],
    permanova_full$R2[which(rownames(permanova_full) == "SampleType")],
    permanova_full$R2[which(rownames(permanova_full) == "Age")]
  ),
  F_statistic = c(
    permanova_full$F[which(rownames(permanova_full) == "Mother")],
    permanova_full$F[which(rownames(permanova_full) == "SampleType")],
    permanova_full$F[which(rownames(permanova_full) == "Age")]
  ),
  p_value = c(
    permanova_full$`Pr(>F)`[which(rownames(permanova_full) == "Mother")],
    permanova_full$`Pr(>F)`[which(rownames(permanova_full) == "SampleType")],
    permanova_full$`Pr(>F)`[which(rownames(permanova_full) == "Age")]
  )
) %>%
  mutate(
    `R² (%)` = round(R2 * 100, 1),
    `F-statistic` = round(F_statistic, 2),
    `p-value` = ifelse(p_value < 0.001, "< 0.001", round(p_value, 3)),
    `Effect Size` = case_when(
      R2 < 0.01 ~ "Negligible",
      R2 < 0.06 ~ "Small",
      R2 < 0.14 ~ "Medium",
      TRUE ~ "Large"
    )
  ) %>%
  dplyr::select(Factor, `R² (%)`, `F-statistic`, `p-value`, `Effect Size`)
knitr::kable(effect_table, format = "html", align = "lcccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, 
                font_size = 22) %>%
  row_spec(1, bold = TRUE, background = "#ffe6e6") %>%  # Highlight Mother
  row_spec(2, bold = TRUE, background = "#e6f3ff")      # Highlight SampleType
```
.footnote[
**PERMANOVA (Bray-Curtis, 9,999 permutations)** - Tests for differences in community composition  
**R²** = Proportion of variance explained | **Cohen's guidelines:** Small (1-6%), Medium (6-14%), Large (>14%)
]
---
## Result 2: Environmental Effects
```{r fig3, fig.height=5.5, fig.width=12}
physeq_subset <- subset_samples(physeq_rarefied, Age %in% c("7+days", "15+days"))
bray_subset <- phyloseq::distance(physeq_subset, method = "bray")
pcoa_subset <- ordinate(physeq_subset, method = "PCoA", distance = bray_subset)
p3 <- plot_ordination(physeq_subset, pcoa_subset, color = "SampleType", shape = "Age") +
  geom_point(size = 6) +
  scale_color_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  labs(title = "PCoA: Muck vs. Sterile") +
  theme(legend.position = "right")
p4 <- ggplot(alpha_data_rarefied %>% filter(Age %in% c("7+days", "15+days")), 
             aes(x = SampleType, y = Shannon, fill = SampleType)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.5) +
  scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 6) +
  labs(title = "Alpha Diversity", y = "Shannon Index") +
  theme(legend.position = "none")
p3 + p4
```
.footnote[**Finding:** Clear separation between Muck and Sterile environments (p < 0.01)]
---
## Taxonomic Composition
```{r fig4, fig.height=6, fig.width=12}
physeq_family <- tax_glom(physeq_rarefied, taxrank = "Family")
physeq_rel_f <- transform_sample_counts(physeq_family, function(x) x / sum(x))
melted_phy_f <- psmelt(physeq_rel_f)
top_families <- names(sort(tapply(melted_phy_f$Abundance, melted_phy_f$Family, sum), decreasing = TRUE))[1:12]
melted_phy_f$Family_Collapsed <- ifelse(melted_phy_f$Family %in% top_families, 
                                         as.character(melted_phy_f$Family), "Other")
ggplot(melted_phy_f, aes(x = Sample, y = Abundance, fill = Family_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ Age, scales = "free_x", space = "free") +
  scale_fill_viridis_d(name = "Family", option = "turbo") +
  labs(y = "Relative Abundance", title = "Family-Level Composition Across Development") +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```
---
## Indicator Species
**Taxa that significantly distinguish Muck from Sterile:**
```{r indicators}
physeq_indic <- subset_samples(physeq_rarefied, Age %in% c("7+days", "15+days", "20+days"))
physeq_indic <- prune_taxa(taxa_sums(physeq_indic) > 0, physeq_indic)
otu_indic <- as(otu_table(physeq_indic), "matrix")
if(taxa_are_rows(physeq_indic)) { otu_indic <- t(otu_indic) }
groups_indic <- sample_data(physeq_indic)$SampleType
indval <- multipatt(otu_indic, groups_indic, control = how(nperm=999))
if (!is.null(indval$sign)) {
  sig_ind <- indval$sign %>% filter(p.value < 0.05)
  if (nrow(sig_ind) > 0) {
    tax_ind <- as.data.frame(tax_table(physeq_indic)[rownames(sig_ind), ])
    tax_ind$Taxon <- paste(tax_ind$Family, tax_ind$Genus, sep = " : ")
    results_ind <- cbind(sig_ind, tax_ind)
    knitr::kable(results_ind[1:min(6, nrow(results_ind)), c("stat", "p.value", "Taxon")], 
                 col.names = c("Indicator Value", "p-value", "Taxon"),
                 digits = 3,
                 format = "html") %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), 
                    full_width = FALSE, 
                    font_size = 20)
  }
}
```
.footnote[**Key Insight:** *Cetobacterium*, *Morganella*, and others are biomarkers]
---
## Core Microbiome (Prevalence > 50%)
```{r core}
get_core_taxa <- function(physeq_obj, threshold = 0.5) {
  physeq_rel <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  core_taxa <- filter_taxa(physeq_rel, function(x) sum(x > 0) > (threshold * length(x)), TRUE)
  return(taxa_names(core_taxa))
}
physeq_muck <- subset_samples(physeq_rarefied, 
                              SampleType == "Muck" & Age %in% c("7+days", "15+days", "20+days"))
physeq_sterile <- subset_samples(physeq_rarefied, 
                                 SampleType == "Sterile" & Age %in% c("7+days", "15+days", "20+days"))
core_muck <- get_core_taxa(physeq_muck)
core_sterile <- get_core_taxa(physeq_sterile)
shared_core <- intersect(core_muck, core_sterile)
unique_muck <- setdiff(core_muck, core_sterile)
unique_sterile <- setdiff(core_sterile, core_muck)
# Create a comprehensive table
all_core <- c(shared_core, unique_muck, unique_sterile)
if(length(all_core) > 0) {
  physeq_core <- prune_taxa(all_core, physeq_rarefied)
  core_tax <- as.data.frame(tax_table(physeq_core))
  core_tax$OTU_ID <- rownames(core_tax)
  core_tax$Status <- case_when(
    rownames(core_tax) %in% shared_core ~ "Shared",
    rownames(core_tax) %in% unique_muck ~ "Muck Only",
    rownames(core_tax) %in% unique_sterile ~ "Sterile Only"
  )
  core_tax$Taxon <- paste(core_tax$Family, core_tax$Genus, sep = " : ")
  
  # Display table
  core_display <- core_tax %>%
    dplyr::select(Status, Taxon) %>%
    arrange(Status)
  
  knitr::kable(core_display, 
               col.names = c("Core Status", "Taxonomic Identity (Family : Genus)"),
               format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), 
                  full_width = FALSE, 
                  font_size = 18) %>%
    row_spec(which(core_display$Status == "Muck Only"), background = "#f4e4d7") %>%
    row_spec(which(core_display$Status == "Shared"), background = "#d4e9d4") %>%
    row_spec(which(core_display$Status == "Sterile Only"), background = "#d9e9e9")
}
```
.footnote[
**Interpretation:** Shared = Maternally transmitted; Unique = Environment-recruited  
**Statistical rigor:** Prevalence threshold = 50% within treatment group (7+, 15+, 20+ days)
]
---
class: center, middle
## Conclusions
---
## Key Findings
.pull-left[
### 1. Maternal Legacy
- Maternal identity is the **primary driver** of early microbiome composition (R² = 20%)
- Strong vertical transmission signal
### 2. Environmental Plasticity
- Exposure to **Muck vs. Sterile** drives divergence in later stages
- Significant effect in 7+ and 15+ day samples
]
.pull-right[
### 3. Specific Drivers
- Identified **4 significant indicator species**
- Defined **core microbiomes** for each environment
- 11 total core taxa across both treatments
### 4. Biological Implication
Early life environment **interacts with** maternal inoculum to shape the adult microbiome
]
---
## Next Steps
**Validation:**
- qPCR confirmation of indicator species
- Shotgun metagenomics for functional potential
**Expanded Analysis:**
- Larger sample size (target n=100)
- Additional time points for temporal dynamics
- Longitudinal tracking of individuals
**Mechanistic Studies:**
- Investigate "Muck-only" core taxa functions
- Test maternal transmission routes (vertical vs. environmental)
---
class: center, middle
# Thank You
**Questions?**
.footnote[
Analysis code available in `Publication_Report.Rmd`  
Figures and tables exported as PDFs and CSVs
]
