---
title: "Microbiome Analysis Results"
author: "Edgardo L. Rosado Ramos"
date: "`r Sys.Date()`"
output: 
  slidy_presentation:
    font_adjustment: +1
---

## Study Overview

- **Objective:** Investigate the effects of **Maternal Identity** vs. **Environmental Exposure** (Muck vs. Sterile) on the developing microbiome.
- **Sample Size:** n = 32 samples across 5 developmental stages.
- **Key Findings:**
    - Strong maternal effect (vertical transmission).
    - Significant environmental effect in later stages.
    - Identification of specific "Core" and "Indicator" taxa.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')
library(phyloseq)
library(vegan)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(kableExtra)
library(DESeq2)
library(indicspecies)
theme_set(theme_bw(base_size = 14))
```

```{r load-data, cache=TRUE}
# Load OTU Table
otu_raw <- read.table("kohler.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t")
otu_mat <- otu_raw[, -(1:3)]
rownames(otu_mat) <- otu_raw$Group 
otu_mat <- t(otu_mat)
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)

# Load Taxonomy
tax_raw <- read.table("kohler.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU

taxonomy_clean <- lapply(strsplit(taxonomy_strings, ";"), function(x) {
  x <- x[nchar(x) > 0]
  x <- gsub("\\(.*?\\)|\\\"", "", x)
  trimws(x)
})

max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, `length<-`, max_ranks))
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)
tax_tab <- tax_table(as.matrix(taxonomy_mat))

# Load Metadata
Kohler.design <- read.table("Kohlerdesign.txt",
                            header = TRUE, sep = "\t", row.names = 1,
                            stringsAsFactors = TRUE, check.names = FALSE)

# Create Phyloseq Object
common_samples <- intersect(sample_names(otu_tab), rownames(Kohler.design))
otu_tab_pruned <- prune_samples(common_samples, otu_tab)
sample_data_obj <- sample_data(Kohler.design[common_samples, ])
physeq <- phyloseq(otu_tab_pruned, tax_tab, sample_data_obj)

# Rarefy
physeq_filt <- filter_taxa(physeq, function(x) sum(x > 3) > (0.2 * length(x)), TRUE)
min_depth <- min(sample_sums(physeq_filt))
set.seed(42)
physeq_rarefied <- rarefy_even_depth(physeq_filt, sample.size = min_depth, replace = FALSE)

# Alpha Diversity
alpha_diversity_rarefied <- estimate_richness(physeq_rarefied, measures = c("Observed", "Shannon"))
sample_metadata_rarefied <- as.data.frame(sample_data(physeq_rarefied))
alpha_data_rarefied <- cbind(alpha_diversity_rarefied, sample_metadata_rarefied)
age_order <- c("Fetus", "Newborn", "7+days", "15+days", "20+days")
alpha_data_rarefied$Age <- factor(alpha_data_rarefied$Age, levels = age_order)
```

## Sample Distribution

```{r fig1, fig.height=4, fig.width=8}
sample_summary <- alpha_data_rarefied %>%
  group_by(Age, SampleType) %>%
  summarise(n = n(), .groups = "drop")

ggplot(sample_summary, aes(x = Age, y = SampleType, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 6, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#d73027", high = "#1a9850", limits = c(0, 10)) +
  labs(title = "Samples per Group", x = "Stage", y = "Treatment") +
  theme_minimal(base_size = 16)
```

**Limitation:** Small sample sizes in some groups (Fetus, Newborn).

## Maternal Clustering

```{r fig2, fig.height=5, fig.width=9}
bray_dist <- phyloseq::distance(physeq_rarefied, method = "bray")
metadata <- data.frame(sample_data(physeq_rarefied))
pcoa_res <- ordinate(physeq_rarefied, method = "PCoA", distance = bray_dist)

plot_ordination(physeq_rarefied, pcoa_res, color = "Mother") +
  geom_point(size = 5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Clustering by Maternal ID (PERMANOVA R² ~ 20%, p < 0.001)")
```

**Finding:** Samples cluster strongly by **Maternal ID**.

## Environmental Effects

```{r fig3, fig.height=5, fig.width=10}
physeq_subset <- subset_samples(physeq_rarefied, Age %in% c("7+days", "15+days"))
bray_subset <- phyloseq::distance(physeq_subset, method = "bray")
pcoa_subset <- ordinate(physeq_subset, method = "PCoA", distance = bray_subset)

p3 <- plot_ordination(physeq_subset, pcoa_subset, color = "SampleType", shape = "Age") +
  geom_point(size = 5) +
  scale_color_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  labs(title = "Muck vs. Sterile (7+ & 15+ Days)")

p4 <- ggplot(alpha_data_rarefied %>% filter(Age %in% c("7+days", "15+days")), 
             aes(x = SampleType, y = Shannon, fill = SampleType)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Muck" = "#8c510a", "Sterile" = "#01665e")) +
  labs(y = "Shannon Index") +
  theme(legend.position = "none")

p3 + p4
```

**Finding:** Clear separation between Muck and Sterile environments.

## Taxonomic Composition

```{r fig4, fig.height=5, fig.width=10}
physeq_family <- tax_glom(physeq_rarefied, taxrank = "Family")
physeq_rel_f <- transform_sample_counts(physeq_family, function(x) x / sum(x))
melted_phy_f <- psmelt(physeq_rel_f)
top_families <- names(sort(tapply(melted_phy_f$Abundance, melted_phy_f$Family, sum), decreasing = TRUE))[1:12]
melted_phy_f$Family_Collapsed <- ifelse(melted_phy_f$Family %in% top_families, as.character(melted_phy_f$Family), "Other")

ggplot(melted_phy_f, aes(x = Sample, y = Abundance, fill = Family_Collapsed)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ Age, scales = "free_x", space = "free") +
  scale_fill_viridis_d(name = "Family") +
  labs(y = "Relative Abundance") +
  theme(axis.text.x = element_blank())
```

## Indicator Species

```{r indicators, results='asis'}
physeq_indic <- subset_samples(physeq_rarefied, Age %in% c("7+days", "15+days", "20+days"))
physeq_indic <- prune_taxa(taxa_sums(physeq_indic) > 0, physeq_indic)
otu_indic <- as(otu_table(physeq_indic), "matrix")
if(taxa_are_rows(physeq_indic)) { otu_indic <- t(otu_indic) }
groups_indic <- sample_data(physeq_indic)$SampleType
indval <- multipatt(otu_indic, groups_indic, control = how(nperm=999))

if (!is.null(indval$sign)) {
  sig_ind <- indval$sign %>% filter(p.value < 0.05)
  if (nrow(sig_ind) > 0) {
    tax_ind <- as.data.frame(tax_table(physeq_indic)[rownames(sig_ind), ])
    tax_ind$Label <- paste(tax_ind$Family, tax_ind$Genus, sep = " : ")
    results_ind <- cbind(sig_ind, tax_ind)
    knitr::kable(results_ind[, c("stat", "p.value", "Label")], 
                 col.names = c("Stat", "p-val", "Taxon"))
  }
}
```

**Key Insight:** Specific bacteria distinguish treatment groups.

## Core Microbiome

```{r core, results='asis'}
get_core_taxa <- function(physeq_obj, threshold = 0.5) {
  physeq_rel <- transform_sample_counts(physeq_obj, function(x) x / sum(x))
  core_taxa <- filter_taxa(physeq_rel, function(x) sum(x > 0) > (threshold * length(x)), TRUE)
  return(taxa_names(core_taxa))
}

physeq_muck <- subset_samples(physeq_rarefied, SampleType == "Muck" & Age %in% c("7+days", "15+days", "20+days"))
physeq_sterile <- subset_samples(physeq_rarefied, SampleType == "Sterile" & Age %in% c("7+days", "15+days", "20+days"))
core_muck <- get_core_taxa(physeq_muck)
core_sterile <- get_core_taxa(physeq_sterile)

shared_core <- intersect(core_muck, core_sterile)
unique_muck <- setdiff(core_muck, core_sterile)
unique_sterile <- setdiff(core_sterile, core_muck)

cat("**Shared Core:**", length(shared_core), "taxa\n\n")
cat("**Muck Only:**", length(unique_muck), "taxa\n\n")
cat("**Sterile Only:**", length(unique_sterile), "taxa")
```

## Conclusions

1. **Maternal Legacy:** Maternal identity drives early microbiome composition
2. **Environmental Plasticity:** Muck vs. Sterile exposure shapes later development
3. **Specific Drivers:** Identified indicator species and core taxa
4. **Implication:** Early environment interacts with maternal inoculum

**Next Steps:** Validate findings with qPCR, investigate functional potential

