---
title: "hawk2"
author: "Edgardo L. Rosado Ramos"
date: "2025-09-16"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
setwd("~/Downloads/Project")

library(phyloseq)
library(dplyr)
library(tidyr)
library(tibble)
library(tidyverse)
```


```{r}
#make phyloseq object
library(phyloseq)
library(microbiome)  # Optional, for further analysis
library(tidyverse)   # For data manipulation

otu_file <- "final.opti_mcc.0.03.subsample.shared"
tax_file <- "final.opti_mcc.0.03.cons.taxonomy"
metadata_file <- "hawkm.design"

# Read OTU table
otu_raw <- read.table("final.opti_mcc.0.03.subsample.shared", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Remove metadata columns
otu_mat <- otu_raw[, -(1:3)]  # removes label, Group, numOtus

# Set rownames as sample IDs
rownames(otu_mat) <- otu_raw$Group

# Transpose so OTUs are rows
otu_mat <- t(otu_mat)

# Create phyloseq OTU table
otu_tab <- otu_table(otu_mat, taxa_are_rows = TRUE)



```

```{r}
# --- 1. Read the taxonomy file ---
tax_raw <- read.table("final.opti_mcc.0.03.cons.taxonomy", 
                      sep = "\t", header = TRUE, stringsAsFactors = FALSE, quote = "")

# --- 2. Extract taxonomy strings and OTU IDs ---
taxonomy_strings <- tax_raw$Taxonomy
names(taxonomy_strings) <- tax_raw$OTU

# --- 3. Split taxonomy strings by ';' and clean ---
taxonomy_clean <- strsplit(taxonomy_strings, ";")

taxonomy_clean <- lapply(taxonomy_clean, function(x) {
  x <- x[nchar(x) > 0]                # Remove empty strings
  x <- gsub("\\(.*?\\)", "", x)       # Remove confidence scores in parentheses
  x <- gsub('\\"', '', x)              # Remove escaped quotes
  trimws(x)                           # Trim whitespace
})

# --- 4. Pad each taxonomy vector to length 7 (Kingdom to Species) ---
max_ranks <- 7
taxonomy_mat <- do.call(rbind, lapply(taxonomy_clean, function(x) {
  length(x) <- max_ranks  # pad with NAs if needed
  x
}))

# --- 5. Assign column and row names ---
colnames(taxonomy_mat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxonomy_mat) <- names(taxonomy_strings)

# --- 6. Create tax_table object for phyloseq ---
tax_tab <- tax_table(as.matrix(taxonomy_mat))

# --- Optional: Check the result ---
head(tax_tab)


```

```{r}
# 1. Extract your taxonomy table into a data frame, which is easy to search
tax_table_df <- as.data.frame(tax_table(physeq_final))

# 2. Search for the specific Genus name "Chlamydophila"
#    This creates a smaller table containing only the matching rows.
chlamydia_result <- subset(tax_table_df, Genus == "Chlamydophila")

# 3. Check the result
if (nrow(chlamydia_result) > 0) {
  
  # 4. Print the full taxonomy line and the OTU ID
  cat("Found a match for Chlamydophila!\n\n")
  print(chlamydia_result)
  
  otu_id_of_chlamydia <- rownames(chlamydia_result)
  cat("\nThe OTU ID is:", otu_id_of_chlamydia, "\n")
  
} else {
  
  cat("Could not find an exact match for the Genus 'Chlamydophila'.\n")
  cat("Checking the Family level ('Chlamydiaceae') instead...\n")
  
  # --- TROUBLESHOOTING: Check the Family level ---
  # Sometimes the classification doesn't reach the Genus level.
  family_result <- subset(tax_table_df, Family == "Chlamydiaceae")
  
  if (nrow(family_result) > 0) {
    cat("\nFound a match at the Family level:\n")
    print(family_result)
    cat("\nThe OTU ID(s) are:", rownames(family_result), "\n")
  } else {
    cat("No matches found at the Genus or Family level.\n")
  }
}
```

```{r}
# Find the OTU ID for Chlamydophila in your tax_table
otu_id_chlamydophila <- "Otu000004" # <-- Replace with the actual OTU ID

# Get the abundance of just that OTU across all samples
chlamydia_abundances <- otu_table(physeq_final)[otu_id_chlamydophila, ]

# Find the sample with the highest abundance
high_abundance_sample <- names(which.max(chlamydia_abundances))
highest_abundance_value <- max(chlamydia_abundances)
sample_metadata <- sample_data(physeq_final)[high_abundance_sample, ]
cat("--- Chlamydophila Abundance Report ---\n")
cat("The Chlamydophila OTU ID is:", otu_id_chlamydoph
ila, "\n")
cat("The highest abundance value is:", highest_abundance_value, "\n")
cat("The sample with the highest Chlamydophila abundance is:", high_abundance_sample, "\n")
cat("\nThe full metadata for this turtle is:\n")
print(sample_metadata)
```

```{r}
# --- Correctly Import the Multi-Column Design File ---

# Define the file path
metadata_file <- "hawkm.design"

# Read the table using arguments that match your new file structure.
# header = TRUE:       CRITICAL. This tells R that your file's first line contains column names.
# row.names = 1:       CRITICAL. This uses the first column ("Group") as the unique row identifier.
# sep = "\t":          IMPORTANT. This tells R that the columns are separated by TABS.
metadata_df <- read.table(metadata_file, 
                          header = TRUE, 
                          row.names = 1, 
                          sep = "\t", 
                          stringsAsFactors = TRUE) # Good practice for categorical data

# Convert the formatted data frame into a phyloseq sample_data object
sample_data_obj <- sample_data(metadata_df)

# --- Verification Step ---
# Always check the imported data to confirm it was read correctly.
# The output should show your new columns: SampleType, Location, and size.
print("--- Verification: First 6 rows of imported metadata ---")
head(sample_data_obj)
```

```{r}
physeq <- phyloseq(otu_tab, tax_tab, sample_data_obj)
physeq

physeq_final <- physeq

```
```{r}
plot_bar(physeq, fill = "Phylum")

physeq_filtered <- filter_taxa(physeq, function(x) sum(x > 0) > 1, prune = TRUE)

plot_richness(physeq, x = "SampleType", measures = c("Shannon", "Observed"))

ord <- ordinate(physeq, method = "NMDS", distance = "bray")
plot_ordination(physeq, ord, color = "SampleType")

library(ggpubr)
```

```{r}
# Extract richness values
richness <- estimate_richness(physeq_no_control, measures = c("Shannon", "Observed"))
richness$SampleType <- sample_data(physeq_no_control)$SampleType

# Boxplot + test
ggboxplot(richness, x = "SampleType", y = "Shannon", add = "jitter") +
  stat_compare_means(method = "wilcox.test")
```

```{r}
library(phyloseq)

# 1. Transform counts to relative abundance
physeq_relabun <- transform_sample_counts(physeq, function(x) x / sum(x))

# 2. Subset control samples
control_samples <- subset_samples(physeq_relabun, SampleType == "Control")

# 3. Extract OTU table from control samples (samples in columns, taxa in rows)
control_otu <- otu_table(control_samples)

# 4. Calculate average abundance of each OTU across control samples
# Convert to matrix/data.frame for easy calculations
control_otu_df <- as.data.frame(as.matrix(control_otu))
avg_control_abund <- rowMeans(control_otu_df)

# 5. Extract full OTU table from all samples
all_otu <- otu_table(physeq_relabun)
all_otu_df <- as.data.frame(as.matrix(all_otu))

# 6. Subtract control average from each sample (make sure no negative values)
corrected_otu_df <- sweep(all_otu_df, 1, avg_control_abund, FUN = "-")
corrected_otu_df[corrected_otu_df < 0] <- 0

# 7. Convert corrected OTU table back to otu_table object
corrected_otu_table <- otu_table(as.matrix(corrected_otu_df), taxa_are_rows = TRUE)

# 8. Subset to remove control samples from original phyloseq object
physeq_no_control <- subset_samples(physeq_relabun, SampleType != "Control")

# 9. Replace OTU table with corrected abundances (only for non-control samples)
# Make sure to keep only samples present in physeq_no_control
corrected_otu_table_no_control <- prune_samples(sample_names(physeq_no_control), corrected_otu_table)

# 10. Create new phyloseq object with corrected OTU table, same sample_data and taxonomy
physeq_corrected <- phyloseq(corrected_otu_table_no_control, 
                            tax_table(physeq_no_control), 
                            sample_data(physeq_no_control))

# Done! You can now analyze physeq_corrected without control samples and corrected for control abundances.


```

```{r}
library(ggplot2)
library(phyloseq)

# Subset raw counts phyloseq to remove controls
physeq_no_control_counts <- subset_samples(physeq, SampleType != "Control")
physeq_no_control_counts <- prune_taxa(taxa_sums(physeq_no_control_counts) > 0, physeq_no_control_counts)

# Now plot richness (counts)
alpha_plot <- plot_richness(physeq_no_control_counts, x = "SampleType", measures = c("Shannon", "Observed")) +
  geom_boxplot(aes(fill = SampleType), alpha = 0.6, outlier.shape = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Alpha Diversity by Sample Type",
       x = "Sample Type",
       y = "Diversity Index") +
  theme_minimal()

print(alpha_plot)

```
```{r}
library(phyloseq)
library(ggplot2)

# Assuming 'physeq' is your full phyloseq object with all samples (including controls)

# Calculate Bray-Curtis distance matrix on full data
bray_dist_all <- distance(physeq, method = "bray")

# Run PCoA ordination on full dataset
pcoa_all <- ordinate(physeq, method = "PCoA", distance = bray_dist_all)

# Extract ordination data for ggplot2
ordination_df_all <- plot_ordination(physeq, pcoa_all, justDF = TRUE)

# Add SampleType metadata
ordination_df_all$SampleType <- sample_data(physeq)$SampleType

# Define colors (add control color if needed)
my_colors_all <- c("Buccal" = "#1b9e77", "Cloaca" = "#d95f02", "Control" = "#7570b3")

# Plot with controls included and colored
ggplot(ordination_df_all, aes(x = Axis.1, y = Axis.2, color = SampleType)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_colors_all) +
  labs(title = "PCoA (Bray-Curtis) including Controls", x = "PC1", y = "PC2") +
  theme_minimal()
```
```{r}
# Calculate Bray-Curtis distance and PCoA ordination
# (Your original code for this part is correct)
bray_dist <- distance(physeq, method = "bray")
pcoa <- ordinate(physeq, method = "PCoA", distance = bray_dist)

# --- CORRECTED PART ---
# Extract PCoA scores for the first two axes.
# We access the ordination results directly from the 'vectors' component of the pcoa object.
ordination_df <- as.data.frame(pcoa$vectors[, 1:2])

# You can now proceed with your analysis using the 'ordination_df' data frame.
# For example, let's look at the first few rows:
head(ordination_df)

```
```{r}
# Title: Publication-Ready PCoA Plot with Professional Defaults
# Description: This script takes a phyloseq object and generates a complete
#              PCoA plot, with pre-selected colors, shapes, and themes.
#
# ACTION REQUIRED:
# 1. Make sure your phyloseq object is named 'mothur_data'.
# 2. Find your metadata column names by running: colnames(sample_data(mothur_data))
# 3. Replace the three placeholder variables in the ggplot section below with your actual column names.
# --- 3. Perform Ordination ---
bray_dist <- distance(physeq, method = "bray")
pcoa <- ordinate(physeq, method = "PCoA", distance = bray_dist)

# --- 4. Prepare Data for Plotting ---
ordination_df <- as.data.frame(pcoa$vectors[, 1:2])
metadata_df <- as(sample_data(physeq), "data.frame")
plot_data <- merge(ordination_df, metadata_df, by = "row.names")
colnames(plot_data)[1] <- "SampleID"

# --- 5. Create the Final PCoA Plot ---
# Calculate the percentage of variance explained
variance_explained <- pcoa$values$Relative_eig[1:2] * 100

# Generate the plot
final_pcoa_plot <- ggplot(plot_data, aes(x = Axis.1, y = Axis.2)) +

  # --- ACTION REQUIRED: EDIT THIS SECTION ---
  # Replace the capitalized placeholder variables with your actual column names.
  geom_point(aes(color = YOUR_PRIMARY_VARIABLE,   # <-- REPLACE THIS
                 shape = YOUR_SECONDARY_VARIABLE), # <-- REPLACE THIS
             size = 4,
             alpha = 0.9) +

  # Add the non-overlapping sample labels
  geom_text_repel(aes(label = SampleID,
                      color = YOUR_LABEL_VARIABLE),    # <-- REPLACE THIS
                  size = 3.5,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  show.legend = FALSE) + # Hide legend for text color
  # --- END OF ACTION REQUIRED SECTION ---

  # Apply a professional, colorblind-friendly color palette
  scale_color_brewer(palette = "Set1") +

  # Manually assign a set of clear, distinct shapes
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 3)) + # Solid circle, triangle, square, diamond, etc.

  # Add plot titles and axis labels
  labs(
    title = "PCoA of Microbial Community Structure",
    subtitle = "Based on Bray-Curtis Dissimilarity",
    x = paste0("Axis 1 (", round(variance_explained[1], 2), "%)"),
    y = paste0("Axis 2 (", round(variance_explained[2], 2), "%)"),
    color = "Primary Group", # Generic legend title
    shape = "Secondary Group" # Generic legend title
  ) +

  # Use a clean, publication-ready theme
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size=5)),
         shape = guide_legend(override.aes = list(size=5))) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  )

# --- 6. Display and Save the Plot ---
print(final_pcoa_plot)

# Save the plot as a high-resolution PNG file
ggsave("Final_PCoA_Plot.png", plot = final_pcoa_plot, width = 11, height = 8.5, dpi = 300)

cat("\nScript finished. Final PCoA plot has been generated and saved as 'Final_PCoA_Plot.png'.\n")

```

```{r}
library(phyloseq)
library(ggplot2)

# Calculate Bray distance matrix
bray_dist <- distance(physeq_corrected, method = "bray")

# Perform PCoA ordination
pcoa <- ordinate(physeq_corrected, method = "PCoA", distance = bray_dist)

# Extract ordination data for ggplot2
ordination_df <- plot_ordination(physeq_corrected, pcoa, justDF = TRUE)
ordination_df$SampleType <- sample_data(physeq_corrected)$SampleType

# Define your custom colors (change these hex codes as you want)
my_colors <- c("Buccal" = "#1b9e77", "Cloaca" = "#d95f02", "Control" = "#7570b3")

# Plot with manual colors
ggplot(ordination_df, aes(x = Axis.1, y = Axis.2, color = SampleType)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_colors) +
  labs(title = "PCoA (Bray-Curtis) of Samples", x = "PC1", y = "PC2") +
  theme_minimal()

```

```{r}
# Agglomerate taxa at Phylum level
physeq_phylum <- tax_glom(physeq_corrected, taxrank = "Phylum")

# Transform to relative abundance for plotting
physeq_phylum_rel <- transform_sample_counts(physeq_phylum, function(x) x / sum(x))

# Plot bar chart
bar_plot <- plot_bar(physeq_phylum_rel, fill = "Phylum") +
  facet_wrap(~ SampleType, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Taxonomic Composition by Phylum")

print(bar_plot)
```

```{r}
plot_bar(physeq_subtracted, fill = "Phylum", facet_grid = ~SampleType)

# Keep only top 20 OTUs by overall abundance (after subtraction)
top_taxa <- names(sort(rowSums(otu_centered), decreasing = TRUE))[1:20]
physeq_top <- prune_taxa(top_taxa, physeq_subtracted)

plot_bar(physeq_top, fill = "Genus", facet_grid = ~SampleType)

```

```{r}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)
library(ComplexHeatmap)
library(DESeq2)
library(rstatix) 


sample_type_cols <- c("Buccal"="#1f78b4", "Cloaca"="#33a02c", "Control"="#e31a1c")
sample_type_shapes <- c("Buccal"=16, "Cloaca"=17, "Control"=15)

theme_set(theme_bw())
```

## Introduction

This report details the analysis of 16S rRNA gene sequencing data from Buccal, Cloaca, and Control samples collected from the Hawksbill Sea Turtle (*Eretmochelys imbricata*).

---

## 1. Alpha Diversity Analysis

Alpha diversity measures the species richness and evenness within a single sample. Here we compare the Shannon diversity index across our three sample types.

```{r alpha-diversity-plot}

alpha_plot <- plot_richness(physeq_final, x = "SampleType", measures = c("Shannon")) +
  geom_boxplot(aes(fill = SampleType), alpha = 0.7) +
  scale_fill_manual(values = sample_type_cols, name = "Sample Type") +
  labs(title = "Alpha Diversity (Shannon Index)", x = "Sample Type", y = "Shannon Diversity") +
  theme(legend.position = "none")

print(alpha_plot)
```

**Statistical Test:** We use the Kruskal-Wallis test to determine if the observed differences in Shannon diversity are statistically significant.

```{r alpha-diversity-stats}
alpha_data <- estimate_richness(physeq_final, measures = "Shannon")
alpha_data$SampleType <- sample_data(physeq_final)$SampleType

kruskal_test_result <- kruskal_test(data = alpha_data, Shannon ~ SampleType)

knitr::kable(kruskal_test_result, caption = "Kruskal-Wallis test for Shannon Diversity")
print(kruskal_test_result)
```

---

## 2. Beta Diversity Analysis

Beta diversity measures the difference in community composition between samples. We use Principal Coordinates Analysis (PCoA) on a Bray-Curtis dissimilarity matrix to visualize these differences.

```{r pcoa-plot}
pcoa_ord <- ordinate(physeq_final, method = "PCoA", distance = "bray")

pcoa_plot <- plot_ordination(physeq_final, pcoa_ord, type = "samples", color = "SampleType", shape = "SampleType") +
  geom_point(size = 4) +
  stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(name = "Sample Type", values = sample_type_cols) +
  scale_shape_manual(name = "Sample Type", values = sample_type_shapes) +
  labs(title = "PCoA of E. imbricata Microbial Communities",
       subtitle = "Bray-Curtis Dissimilarity")

print(pcoa_plot)
```

**Statistical Test:** We use a PERMANOVA test (`adonis2`) to determine if the clustering of sample groups is statistically significant.

```{r beta-diversity-stats, results='asis'}
bray_dist <- phyloseq::distance(physeq_final, method = "bray")
sample_df <- data.frame(sample_data(physeq_final))

permanova_result <- adonis2(bray_dist ~ SampleType, data = sample_df, permutations = 999)

knitr::kable(permanova_result, caption = "PERMANOVA results for Beta Diversity")
print(permanova_result)
```

---

## 3. Taxonomic Composition

### Relative Abundance Bar Plot
This plot shows the relative abundance of the major bacterial Phyla across each sample.

```{r bar-plot}
physeq_phylum <- tax_glom(physeq_final, taxrank = "Phylum")

bar_plot <- plot_bar(physeq_phylum, x = "SampleID", fill = "Phylum") +
  facet_wrap(~SampleType, scales = "free_x", nrow = 1) +
  labs(title = "Relative Abundance by Phylum", x = "Sample", y = "Relative Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8))

print(bar_plot)
```

### Heatmap of Abundant Genera
This heatmap shows the relative abundance of the top 30 most abundant genera.

```{r heatmap}
physeq_genus <- tax_glom(physeq_final, taxrank = "Genus", NArm = FALSE)

top30_taxa <- names(sort(taxa_sums(physeq_genus), decreasing = TRUE)[1:30])
physeq_top30 <- prune_taxa(top30_taxa, physeq_genus)

plot_heatmap(physeq_top30, 
             sample.label = "SampleType", 
             taxa.label = "Genus", 
             sample.order = "SampleType",
             low = "#FFFFFF", high = "#0000FF", na.value = "white",
             title = "Heatmap of Top 30 Most Abundant Genera")
```
```{r}
# Find the OTU ID for Chlamydophila in your tax_table
otu_id_chlamydophila <- "Otu000004" # <-- Replace with the actual OTU ID

# Get the abundance of just that OTU across all samples
chlamydia_abundances <- otu_table(physeq_final)[otu_id_chlamydophila, ]

# Find the sample with the highest abundance
high_abundance_sample <- names(which.max(chlamydia_abundances))

# Get the full metadata for that specific sample
sample_metadata <- sample_data(physeq_final)[high_abundance_sample, ]

cat("The sample with the highest Chlamydophila abundance is:", high_abundance_sample, "\n")
cat("\nThe full metadata for this turtle is:\n")
print(sample_metadata)
```

---

## 4. Differential Abundance Analysis (DESeq2)
To identify which specific taxa are significantly different between groups, we use DESeq2.
**Note:** DESeq2 should ideally be run on non-rarefied data. This example uses your rarefied data for demonstration.

```{r deseq2-analysis}
# Add 1 to all counts to handle zeros, a requirement for DESeq2
physeq_deseq <- physeq_final
otu_table(physeq_deseq) <- otu_table(physeq_deseq) + 1

# Convert phyloseq object to DESeq2 object
deseq_obj <- phyloseq_to_deseq2(physeq_deseq, design = ~ SampleType)

# Run DESeq2
deseq_results <- DESeq(deseq_obj, test = "Wald", fitType = "parametric")

# --- Compare Buccal vs. Control ---
res_buccal_vs_control <- results(deseq_results, contrast = c("SampleType", "Buccal", "Control"))
res_buccal_vs_control_sig <- subset(res_buccal_vs_control, padj < 0.05)
res_buccal_vs_control_sig <- cbind(as.data.frame(res_buccal_vs_control_sig), as.data.frame(tax_table(physeq_final)[rownames(res_buccal_vs_control_sig), ]))

# --- Compare Cloaca vs. Control ---
res_cloaca_vs_control <- results(deseq_results, contrast = c("SampleType", "Cloaca", "Control"))
res_cloaca_vs_control_sig <- subset(res_cloaca_vs_control, padj < 0.05)
res_cloaca_vs_control_sig <- cbind(as.data.frame(res_cloaca_vs_control_sig), as.data.frame(tax_table(physeq_final)[rownames(res_cloaca_vs_control_sig), ]))

print("--- DESeq2 Analysis Completed ---")
```

### Table: Taxa Significantly Different Between Buccal and Control
```{r deseq2-table-1, results='asis'}
knitr::kable(res_buccal_vs_control_sig, caption = "Differentially abundant taxa (p-adj < 0.05) between Buccal and Control samples.")

print(res_buccal_vs_control_sig)

```

### Table: Taxa Significantly Different Between Cloaca and Control
```{r deseq2-table-2, results='asis'}
knitr::kable(res_cloaca_vs_control_sig, caption = "Differentially abundant taxa (p-adj < 0.05) between Cloaca and Control samples.")
print(res_cloaca_vs_control_sig)

```
```{r}
# --- Corrected Bray-Curtis Calculation ---

# It's good practice to normalize to relative abundance before calculating Bray-Curtis
physeq_rel_abund <- transform_sample_counts(physeq, function(x) x/sum(x))

# The corrected function call with 'phyloseq::'
# This tells R to use the specific version of the distance function from the phyloseq package
bray_dist_all <- phyloseq::distance(physeq_rel_abund, method = "bray")

# Now you can proceed with your ordination
pcoa_all <- ordinate(physeq_rel_abund, method = "PCoA", distance = bray_dist_all)

# And plot the result
plot_ordination(physeq_rel_abund, pcoa_all, color = "SampleType") +
  geom_point(size = 4, alpha = 0.7) +
  theme_minimal() +
  labs(title = "PCoA of All Samples (including Controls)")
```

```{r}
# --- PCoA Plot Colored by Location ---
# We use the 'phyloseq_nocontrol' object for biological comparisons.
physeq_rel_abund <- transform_sample_counts(phyloseq_nocontrol, function(x) x / sum(x))
bray_dist <- phyloseq::distance(physeq_rel_abund, method = "bray")
pcoa_ord <- ordinate(physeq_rel_abund, method = "PCoA", distance = bray_dist)

# 'plot_ordination' automatically finds the new metadata columns.
plot_ordination(physeq_rel_abund, pcoa_ord, color = "Location", shape = "SampleType") +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCoA Plot Colored by Location")

# --- Statistical Test (PERMANOVA) ---
# This gives you a p-value for the 'Location' variable.
sample_data_for_test <- as(sample_data(phyloseq_nocontrol), "data.frame")
permanova_location_test <- adonis2(bray_dist ~ Location, data = sample_data_for_test)

print("--- PERMANOVA Test for Location ---")
print(permanova_location_test)
```
